---
title: "Coral Growth — Size-Dependent Growth Rate Analysis"
subtitle: "GLMM analysis of proportional growth rates by taxa, site, habitat, and bleaching period"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
editor: visual
---

# Overview

This document models **size-dependent proportional growth rates** for three coral taxa (*Acropora*, *Pocillopora*, *Porites*) using generalized linear mixed models (GLMMs) fitted with `glmmTMB`.

Two research questions are addressed:

1.  **Model 1 (Habitat):** Does size-dependent growth rate differ between forereef and backreef habitats around the island before and after the 2019 bleaching event?
2.  **Model 2 (Site × Habitat):** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 backreef, LTER1 forereef, LTER2 backreef, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Reference group for all models:** backreef (Model 1) or LTER1 backreef (Model 2), before bleaching.

All models use `log(size_t1)` as the response variable, with `log(size_t0)` as a covariate. The coefficient on `log_size_t0` captures the size-dependence of growth.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(glmmTMB)
```

------------------------------------------------------------------------

# Data Preparation

## Load & Clean

```{r}
#| label: load-clean
#| message: false
#| warning: false

clean_coral <- read_csv(here::here("..", "eds-coral-data-storage-management", "data", "volume_corals.csv"))

# Coral color way - consistent across all visualizations
coral_colors <- c(
  Acr = "blue",
  Poc = "red",
  Por = "darkgreen",
  Mil = "darkorange"
)

clean_coral <- clean_coral |>
  # Removing plots from backreef surveys
  filter(str_detect(transect, "T")) |>
  # Change volume variables to numeric
  mutate(
    height = as.numeric(height),
    width  = as.numeric(width),
    length = as.numeric(length),
    # Create z (m) and y (m) columns
    y_m  = y / 100,
    z_m  = z / 100,
    # Acropora were input as "A" in 2024, converting to "Acr"
    taxa = case_when(taxa == "A" ~ "Acr",
                     TRUE ~ taxa)
  )
```

## Volume Functions

```{r}
#| label: volume-functions

# Sphere volume — used for Pocillopora
sph_vol <- function(l, w, h) {
  r <- (l + w + h) / 3 / 2
  v <- (4/3) * pi * r^3
  return(v)
}

# Hemisphere volume — used for Porites
hemi_sph_vol <- function(l, w, h) {
  r <- (l + w + h) / 3 / 2
  v <- (2/3) * pi * r^3
  return(v)
}

# Cylinder volume — used for Acropora and Millepora
cyl_vol <- function(l, w, h) {
  r <- ((l + w) / 2) / 2
  v <- pi * r^2 * h
  return(v)
}
```

## Calculate Volume

```{r}
#| label: calculate-volume

# Calculate volume per taxa using appropriate geometry
coral_volume_poc <- clean_coral |>
  filter(taxa == "Poc") |>
  rowwise() |>
  mutate(volume = sph_vol(length, width, height)) |>
  ungroup()

coral_volume_por <- clean_coral |>
  filter(taxa == "Por") |>
  rowwise() |>
  mutate(volume = hemi_sph_vol(length, width, height)) |>
  ungroup()

coral_volume <- clean_coral |>
  filter(!taxa %in% c("Poc", "Por")) |>
  rowwise() |>
  mutate(volume = cyl_vol(length, width, height)) |>
  ungroup()

coral_volume_mod <- bind_rows(coral_volume_poc, coral_volume_por, coral_volume)
```

## Create Lagged Size Variables

```{r}
#| label: lag-variables

# Create t0, t1, and t-1 size variables per coral
# period: 0 = before bleaching (<=2018), 1 = after bleaching (>2018)
volume_coral <- coral_volume_mod |>
  arrange(coral_number, year) |>
  group_by(coral_number) |>
  mutate(
    size_t_minus1 = lag(volume),
    size_t0       = volume,
    size_t1       = lead(volume),
    # Year of the lead observation — used to check consecutive years
    year_t1       = lead(year),
    period = factor(case_when(
      year <= 2018 ~ 0,
      year >  2018 ~ 1
    ))
  ) |>
  ungroup()
```

## Filter & Model Preparation

```{r}
#| label: filter-data

# Exclusion criteria:
# - Millepora excluded (insufficient sample size)
# - Require both size measurements (t0 and t1 non-missing)
# - Only consecutive year pairs (year_t1 - year == 1)
# - Remove zero/negative volumes (measurement errors)
# - Remove implausible growth ratios: >10x or <0.1x annual change
#   (confirmed as measurement error — spread across all years/taxa,
#    not concentrated around 2019 bleaching event)
growth_data <- volume_coral |>
  filter(
    taxa != "Mil",
    !is.na(size_t0), !is.na(size_t1),
    !is.na(year_t1), year_t1 - year == 1,  # consecutive years only
    size_t0 > 0, size_t1 > 0,
    size_t1 / size_t0 <= 10,               # remove implausible growth
    size_t1 / size_t0 >= 0.1               # remove implausible shrinkage
  ) |>
  mutate(
    # Unique transect ID — transect numbers repeat across site/habitat combos
    transect_id = paste(site, habitat, transect, sep = "_"),
    # Set reference levels
    taxa        = relevel(factor(taxa),    ref = "Acr"),
    habitat     = relevel(factor(habitat), ref = "BR"),
    site        = relevel(factor(site),    ref = "LTER1"),
    # Log-transform size variables
    log10_size_t0 = log10(size_t0),
    log10_size_t1 = log10(size_t1)
  )

# Split by taxa for separate models
growth_acr <- growth_data |> filter(taxa == "Acr")
growth_poc <- growth_data |> filter(taxa == "Poc")
growth_por <- growth_data |> filter(taxa == "Por")
```

------------------------------------------------------------------------

# Model 1: Habitat-Level Growth Before/After Bleaching

**Question:** Does size-dependent growth rate differ between forereef and backreef habitats before and after the 2019 bleaching event?

**Fixed effects:** `log_size_t0 * habitat * period`

**Random effects:** Colonies nested in transects nested in sites `(1 | site/transect_id/coral_number)` + inter-annual variation `(1 | year)`

**Reference group:** Backreef, before bleaching.

## Fit Models

```{r}
#| label: m1-fit
#| message: false
#| warning: false

m1_acr <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | site/transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_acr
)

m1_poc <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | site/transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_poc
)

m1_por <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | site/transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_por
)
```

## Model Summaries

```{r}
#| label: m1-summaries

summary(m1_acr)
summary(m1_poc)
summary(m1_por)
```

------------------------------------------------------------------------

```{r}
growth_acr$pred_log10_t1 <- predict(m1_acr, type = "response", re.form = NA)

growth_poc$pred_log10_t1 <- predict(m1_poc, type = "response", re.form = NA)

growth_por$pred_log10_t1 <- predict(m1_por, type = "response", re.form = NA)

```

```{r}
 ggplot(data = growth_acr) +
  geom_point(aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3) +
  geom_line(aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                   labels = c("0" = "Before bleaching", "1" = "After bleaching")) + 
  theme_bw() +
  labs(title = "Acropora Growth",
       x = "Initial size (cm³)",
       y = "Final size (cm³)", 
       color = "Period")
 
 ggplot(data = growth_poc) +
  geom_point(aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3) +
  geom_line(aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                   labels = c("0" = "Before bleaching", "1" = "After bleaching")) + 
  theme_bw() +
  labs(title = "Pocillopora Growth",
       x = "Initial size (cm³)",
       y = "Final size (cm³)", 
       color = "Period")
 
 
 ggplot(data = growth_por) +
  geom_point(aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3) +
  geom_line(aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                   labels = c("0" = "Before bleaching", "1" = "After bleaching")) + 
  theme_bw() +
  labs(title = "Porites Growth",
       x = "Initial size (cm³)",
       y = "Final size (cm³)", 
       color = "Period")

   
```

# Model 2: Island-Wide Growth by Site × Habitat Before/After Bleaching

**Question:** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 backreef, LTER1 forereef, LTER2 backreef, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Fixed effects:** `log_size_t0 * site * habitat * period`

**Random effects:** Colonies nested in transects `(1 | transect_id/coral_number)` + inter-annual variation `(1 | year)`

Note: Site is included as a **fixed effect** here because the question explicitly compares specific named sites. Transect nesting is simplified to `transect_id/coral_number` since site is already captured in the fixed effects and in `transect_id`.

**Reference group:** LTER1 backreef, before bleaching.

## Fit Models

```{r}
#| label: m2-fit
#| message: false
#| warning: false

m2_acr <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_acr
)

m2_poc <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_poc
)

m2_por <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_por
)
```

## Model Summaries

```{r}
#| label: m2-summaries

summary(m2_acr)
summary(m2_poc)
summary(m2_por)
```

```{r}
growth_acr$pred_log10_t1 <- predict(m2_acr, type = "response", re.form = NA)

growth_poc$predlog10_t1 <- predict(m2_poc, type = "response", re.form = NA)

growth_por$predlog10_t0 <- predict(m2_por, type = "response", re.form = NA)
```

```{r}
# Create prediction grid
acr_pred_grid <- growth_acr |>
  distinct(site, habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_acr$log10_size_t0),
                              max(growth_acr$log10_size_t0),
                              length.out = 100))

# Generate predictions (population-level, no random effects)
acr_pred_grid$pred_size_t0 <- predict(m2_acr, newdata = acr_pred_grid, 
                          type = "response", re.form = NA)

# Plot
ggplot() +
  geom_point(data = growth_acr, 
             aes(x = log10_size_t0, y = log10_size_t1), 
             alpha = 0.3, size = 0.8) +
  geom_line(data = acr_pred_grid,
            aes(x = log10_size_t0, y = pred_size_t0, color = period)) +
  scale_color_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Before bleaching", "1" = "After bleaching")
  ) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  labs(x = "Initial size (cm³)", y = "Final size (cm³)",
       title = "Acropora Growth", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()
```

```{r}
# Create prediction grid
poc_pred_grid <- growth_poc |>
  distinct(site, habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_poc$log10_size_t0),
                              max(growth_poc$log10_size_t0),
                              length.out = 100))

# Generate predictions (population-level, no random effects)
poc_pred_grid$pred_size_t0 <- predict(m2_poc, newdata = poc_pred_grid, 
                          type = "response", re.form = NA)

# Plot
ggplot() +
  geom_point(data = growth_poc, 
             aes(x = log10_size_t0, y = log10_size_t1), 
             alpha = 0.3, size = 0.8) +
  geom_line(data = poc_pred_grid,
            aes(x = log10_size_t0, y = pred_size_t0, color = period)) +
  scale_color_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Before bleaching", "1" = "After bleaching")
  ) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  labs(x = "Initial size (cm³)", y = "Final size (cm³)",
       title = "Pocillopora Growth", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()
```

```{r}
# Create prediction grid
por_pred_grid <- growth_por |>
  distinct(site, habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_por$log10_size_t0),
                              max(growth_por$log10_size_t0),
                              length.out = 100))

# Generate predictions (population-level, no random effects)
por_pred_grid$pred_size_t0 <- predict(m2_por, newdata = por_pred_grid, 
                          type = "response", re.form = NA)

# Plot
ggplot() +
  geom_point(data = growth_por, 
             aes(x = log10_size_t0, y = log10_size_t1), 
             alpha = 0.3, size = 0.8) +
  geom_line(data = por_pred_grid,
            aes(x = log10_size_t0, y = pred_size_t0, color = period)) +
  scale_color_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Before bleaching", "1" = "After bleaching")
  ) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  labs(x = "Initial size (cm³)", y = "Final size (cm³)",
       title = "Porites Growth", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()
```
