---
title: "Survivorship GLMM"
format: html
execute:
  warning: false
  
---

# Overview

This document models **size-dependent survivorship** for three coral taxa (*Acropora*, *Pocillopora*, *Porites*) using generalized linear mixed models (GLMMs) fitted with `glmmTMB`.

Two research questions are addressed:

1.  **Model 1 (Habitat):** Does size-dependent survival probability differ between forereef and lagoon habitats around the island before and after the 2019 bleaching event?
2.  **Model 2 (Site × Habitat):** Does size-dependent survival probability differ across all site-habitat combinations (LTER1 lagoon, LTER1 forereef, LTER2 lagoon, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Reference group for all models:** lagoon (Model 1) or LTER1 lagoon (Model 2), before bleaching.

# Setup

```{r}
library(tidyverse)
library(here)
library(MASS)
library(glmmTMB)
library(performance)
library(DHARMa)
library(emmeans)
library(dplyr)
library(purrr)
library(gt)
```

# Data Preparation
```{r}
coral <- read_csv(here::here("..", "eds-coral-data-storage-management", "analysisdata", "coral_demography_analysis_data.csv"), guess_max = 10000) # Corrects column types

coral_clean <- coral %>% 
  group_by(coral_id) %>%
  arrange(coral_id, year) %>% 
  # Assign previous year's size, when a dead coral does not have a current measurement
  mutate(current_volume = case_when(
    is.na(size_t0) ~ size_t_minus1,
    TRUE~ size_t0
  ), # Create a survival variable
  survival = case_when(
    dyn_death == 1 ~ 0, # If a coral dies assigned a 0
    is.na(l) ~ NA, # All non existent corals are excluded
    TRUE ~ 1 # All living corals assigned 1
  ), 
  # Create bleaching event variable
  period = as_factor(case_when( 
    year < 2019 ~ 0,
    TRUE ~ 1
  )),
  # Transform size
  log_size = log10(current_volume)) %>% 
  ungroup() %>%
  filter(!is.na(survival))


vars_used <- c("survival", "log_size", "site", "habitat", "period", 
               "transect_id", "coral_id", "year") # Variables for GLMMs

coral_death <- coral_clean[complete.cases(coral_clean[, vars_used]), ] # Select only corals with all fixed and random variables
```

# Model 1: Habitat-Level Growth Before/After Bleaching

**Question:** Does size-dependent growth rate differ between forereef and lagoon habitats before and after the 2019 bleaching event?

**Fixed effects:** `log_size_t0 * habitat * period`

**Random effects:** Colonies nested in transects nested in sites `(1 | site/transect_id/coral_id)` + inter-annual variation `(1 | year)`

**Reference group:** lagoon, before bleaching.

## Fit models

### *Acropora*
```{r}
coral_acr_hab <- coral_death %>% 
  filter(taxa == "Acr")

acr_hab_surv_mod <- glmmTMB(survival ~ log_size * habitat * period + (1 | site/transect_id/coral_id) + (1| year), family=binomial, data = coral_acr_hab, na.action = na.exclude)

coral_acr_hab$pred_surv <- predict(acr_hab_surv_mod, type = "response", re.form = NA)

```

### *Pocillilopora*
```{r}
coral_poc_hab <- coral_death %>% 
  filter(taxa == "Poc")

poc_hab_surv_mod <- glmmTMB(survival ~ log_size * habitat * period + (1 | site/transect_id/coral_id) + (1| year), family=binomial, data = coral_poc_hab, na.action = na.exclude)

coral_poc_hab$pred_surv <- predict(poc_hab_surv_mod, type = "response", re.form = NA)
```

### *Porites*
```{r}
coral_por_hab <- coral_death %>% 
  filter(taxa == "Por")

por_hab_surv_mod <- glmmTMB(survival ~ log_size * habitat * period + (1 | site/transect_id/coral_id) + (1| year), family=binomial, data = coral_por_hab, na.action = na.exclude)

coral_por_hab$pred_surv <- predict(por_hab_surv_mod, type = "response", re.form = NA)
```

#### Model Summaries
```{r}
# Acropora habitat
summary(acr_hab_surv_mod)

# Pocillopora habitat
summary(poc_hab_surv_mod)

# Porites habitat
summary(por_hab_surv_mod)
```

## EEMeans
```{r}
# Acropora
emtrends(acr_surv_mod, ~ site * habitat * period, var = "log_size")
pairs(emtrends(acr_surv_mod, ~ site * habitat * period, var = "log_size"))

# Pocillopora
emtrends(poc_surv_mod, ~ site * habitat * period, var = "log_size")
pairs(emtrends(poc_surv_mod, ~ site * habitat * period, var = "log_size"))

# Porites
emtrends(por_surv_mod, ~ site * habitat * period, var = "log_size")
pairs(emtrends(por_surv_mod, ~ site * habitat * period, var = "log_size"))
```

## Constrast tables
```{r}

models <- list(ACR = acr_hab_surv_mod, POC = poc_hab_surv_mod, POR = por_hab_surv_mod)

# Group-specific slopes
slopes <- imap_dfr(models, ~ as.data.frame(
  emtrends(.x, ~ habitat * period, var = "log_size")
) %>% mutate(taxon = .y))

# Within-habitat period contrasts
contrasts <- imap_dfr(models, ~ as.data.frame(
  pairs(emtrends(.x, ~ period | habitat, var = "log_size"))
) %>% mutate(taxon = .y))

# Clean table
contrasts %>%
  dplyr::select(taxon, habitat, contrast, estimate, SE, z.ratio, p.value) %>%
  mutate(
    across(c(estimate, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) %>%
  arrange(taxon, habitat) %>%
  gt(groupname_col = "taxon") %>%
  cols_label(
    habitat = "Habitat",
    contrast = "Contrast",
    estimate = "Δ slope",
    z.ratio = "z",
    p.value = "p",
    sig = ""
  ) %>%
  tab_header(title = "Change in size-dependent growth slope after 2019 bleaching")
```

## Visualize models
### *Acropora*
```{r}
ggplot(coral_acr_hab, aes(log_size, survival)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_line(aes(y = pred_surv, color = period)) +
  scale_color_manual(values = c("dodgerblue", "red"), labels = c("Pre-bleaching", "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  facet_grid(~habitat, scales = "free", labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  theme_bw() +
  labs(x = "Volume (log₁₀ cm³)",
       y = "Survival probability",
       color = "Period") +
  theme(
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x =  element_text(size = 12),
        strip.text = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
        )

#ggsave(here::here("survival", "figs", "coral_acr_hab_fig.png"), height = 5, width = 10)
```

### *Pocillopora*
```{r}
ggplot(coral_poc_hab, aes(log_size, survival)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_line(aes(y = pred_surv, color = period)) +
  scale_color_manual(values = c("dodgerblue", "red"), labels = c("Pre-bleaching", "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  facet_grid(~habitat, scales = "free", labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  theme_bw() +
  labs(x = "Volume (log₁₀ cm³)",
       y = "Survival probability",
       color = "Period") +
  theme(
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x =  element_text(size = 12),
        strip.text = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
        )

#ggsave(here::here("survival", "figs", "coral_poc_hab_fig.png"), height = 5, width = 10)
```

### *Porites*
```{r}
ggplot(coral_por_hab, aes(log_size, survival)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_line(aes(y = pred_surv, color = period)) +
  scale_color_manual(values = c("dodgerblue", "red"), labels = c("Pre-bleaching", "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  facet_grid(~habitat, scales = "free", labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  theme_bw() +
  labs(x = "Volume (log₁₀ cm³)",
       y = "Survival probability",
       color = "Period") +
  theme(
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x =  element_text(size = 12),
        strip.text = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
        )

#ggsave(here::here("survival", "figs", "coral_por_hab_fig.png"), height = 5, width = 10)
```

## Model checks
### *Acropora* residuals
```{r}
acr_hab_surv_mod_output <- simulateResiduals(fittedModel = acr_hab_surv_mod, plot = FALSE)
plot(acr_hab_surv_mod_output)
```


### *Pocillopora* residuals
```{r}
poc_hab_surv_mod_output <- simulateResiduals(fittedModel = poc_hab_surv_mod, plot = FALSE)
plot(poc_hab_surv_mod_output)
```
### *Porites* residuals
```{r}
por_hab_surv_mod_output <- simulateResiduals(fittedModel = por_hab_surv_mod, plot = FALSE)
plot(por_hab_surv_mod_output)
```

# Model 2: Island-Wide Growth by Site × Habitat Before/After Bleaching

**Question:** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 lagoon, LTER1 forereef, LTER2 lagoon, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Fixed effects:** `log_size_t0 * site * habitat * period`

**Random effects:** Colonies nested in transects `(1 | transect_id/coral_id)` + inter-annual variation `(1 | year)`

Note: Site is included as a **fixed effect** here because the question explicitly compares specific named sites. Transect nesting is simplified to `transect_id/coral_id` since site is already captured in the fixed effects and in `transect_id`.

**Reference group:** LTER1 lagoon, before bleaching.

## Fit models

### *Acropora*
```{r}
coral_acr <- coral_death %>% 
  filter(taxa == "Acr")

acr_surv_mod <- glmmTMB(survival ~ log_size * site * habitat * period + (1 | transect_id/coral_id) + (1| year), family=binomial, data = coral_acr, na.action = na.exclude)

coral_acr$pred_surv <- predict(acr_surv_mod, type = "response", re.form = NA)
```

### *Pocillopera*
```{r}
coral_poc <- coral_death %>% 
  filter(taxa == "Poc")

poc_surv_mod <- glmmTMB(survival ~ log_size * site * habitat * period + (1 | transect_id/coral_id) + (1| year), family=binomial, data = coral_poc, na.action = na.exclude)

coral_poc$pred_surv <- predict(poc_surv_mod, type = "response", re.form = NA)
```

### *Porites*
```{r}
coral_por <- coral_death %>% 
  filter(taxa == "Por")

por_surv_mod <- glmmTMB(survival ~ log_size * site * habitat * period + (1 | transect_id/coral_id) + (1| year), family=binomial, data = coral_por, na.action = na.exclude)

coral_por$pred_surv <- predict(por_surv_mod, type = "response", re.form = NA, na.action = na.exclude)
```

## Model summaries
```{r}
summary(acr_surv_mod)
summary(poc_surv_mod)
summary(por_surv_mod)
```

## EEMeans
```{r}
# Acropora
emtrends(acr_hab_surv_mod, ~ habitat * period, var = "log_size")
pairs(emtrends(acr_hab_surv_mod, ~ habitat * period, var = "log_size"))

# Pocillopora
emtrends(poc_hab_surv_mod, ~ habitat * period, var = "log_size")
pairs(emtrends(poc_hab_surv_mod, ~ habitat * period, var = "log_size"))

# Porites
emtrends(por_hab_surv_mod, ~ habitat * period, var = "log_size")
pairs(emtrends(por_hab_surv_mod, ~ habitat * period, var = "log_size"))
```

## Constrast tables
```{r}

models2 <- list(ACR = acr_surv_mod, POC = poc_surv_mod, POR = por_surv_mod)

# Group-specific slopes
slopes <- imap_dfr(models2, ~ as.data.frame(
  emtrends(.x, ~ site * habitat * period, var = "log_size")
) %>% mutate(taxon = .y))

# Within-habitat period contrasts
contrasts <- imap_dfr(models2, ~ as.data.frame(
  pairs(emtrends(.x, ~ period | site * habitat, var = "log_size"))
) %>% mutate(taxon = .y))

# Clean table
contrasts %>%
  dplyr::select(taxon, site, habitat, contrast, estimate, SE, z.ratio, p.value) %>%
  mutate(
    across(c(estimate, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) %>%
  arrange(taxon, site, habitat) %>%
  gt(groupname_col = "taxon") %>%
  cols_label(
    habitat = "Habitat",
    contrast = "Contrast",
    estimate = "Δ slope",
    z.ratio = "z",
    p.value = "p",
    sig = ""
  ) %>%
  tab_header(title = "Change in size-dependent growth slope after 2019 bleaching")
```


# Visualize models
```{r}
ggplot(coral_acr, aes(log_size, survival)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_line(aes(y = pred_surv, color = period)) +
  scale_color_manual(values = c("dodgerblue", "red"), labels = c("Pre-bleaching", "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  facet_grid(habitat~site, scales = "free", labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  theme_bw() +
  labs(x = "Volume (log₁₀ cm³)",
       y = "Survival probability",
       color = "Period") +
  theme(
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x =  element_text(size = 12),
        strip.text = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
        )

#ggsave(here::here("survival", "figs", "coral_acr_fig.png"), height = 5, width = 10)
```


```{r}
ggplot(coral_poc, aes(log_size, survival)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_line(aes(y = pred_surv, color = period)) +
  scale_color_manual(values = c("dodgerblue", "red"), labels = c("Pre-bleaching", "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  facet_grid(habitat~site, scales = "free", labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  theme_bw() +
  labs(x = "Volume (log₁₀ cm³)",
       y = "Survival probability",
       color = "Period") +
  theme(
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x =  element_text(size = 12),
        strip.text = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
        )

#ggsave(here::here("survival", "figs", "coral_poc_fig.png"), height = 5, width = 10)
```


```{r}
ggplot(coral_por, aes(log_size, survival)) +
  geom_point(alpha = 0.3, color = "black") +
  geom_line(aes(y = pred_surv, color = period)) +
  scale_color_manual(values = c("dodgerblue", "red"), labels = c("Pre-bleaching", "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  facet_grid(habitat~site, scales = "free", labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  theme_bw() +
  labs(x = "Volume (log₁₀ cm³)",
       y = "Survival probability",
       color = "Period") +
  theme(
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x =  element_text(size = 12),
        strip.text = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
        )

#ggsave(here::here("survival", "figs", "coral_por_fig.png"), height = 5, width = 10)
```


## Model Checks

### *Acropora* residuals
```{r}
acr_surv_mod_output <- simulateResiduals(fittedModel = acr_surv_mod, plot = FALSE)
plot(acr_surv_mod_output)
```

### *Pocillopora* residuals
```{r}
poc_surv_mod_output <- simulateResiduals(fittedModel = poc_surv_mod, plot = FALSE)
plot(poc_surv_mod_output)
```

### *Porites* residuals
```{r}
por_surv_mod_output <- simulateResiduals(fittedModel = por_surv_mod, plot = FALSE)
plot(por_surv_mod_output)
```

