---
title: "recruitment_glmm_walkthrough"
format: html
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora from 2013-2024?**

## Data Preparation

### Load and Clean

The following packages were required prior to conducting additional data wrangling:

```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation 
```

Then, the cleaned data (referenced from a separate repository) was loaded in.

```{r}
clean_coral <- read_csv(
  "~/MEDS/capstone/eds-coral-data-storage-management/analysisdata/coral_demography_analysis_data.csv", guess_max = 1000
  # coral_demography_analysis_data.csv
)
str(clean_coral)
```

Standardize recruitment for every 5 meters squared.

```{r}
# Summarize standard recruits to site-habitat-transect and year level 
recruit_transect <- clean_coral %>%
  group_by(taxa, site, habitat, transect, year) %>%
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    recruits_std = n_recruits / 5,
    recruits_int = as.integer(round(recruits_std)),
    pre_post_f   = factor(
      case_when(
        year >= 2013 & year <= 2018 ~ "Pre",
        year >= 2019 & year <= 2024 ~ "Post"
      ),
      levels = c("Pre", "Post")
    ),
    transect_id = paste(site, transect, sep = "_"),
    taxa        = factor(taxa),
    site        = relevel(factor(site),    ref = "LTER1"),
    habitat     = relevel(factor(habitat), ref = "BR")
  ) 
# %>%
#   filter(!is.na(pre_post_f))

view(recruit_transect)
# No NAs
```

```{r}
# Create filtered datasets for each taxa
dat_acr <- recruit_transect %>% filter(taxa == "Acr")
dat_poc <- recruit_transect %>% filter(taxa == "Poc")
dat_por <- recruit_transect %>% filter(taxa == "Por")
```

```{r}
# Exploratory bar chart 
recruit_transect %>%
  filter(taxa != "Mil") %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = pre_post_f)) +
  geom_col() +
  facet_grid(habitat ~ taxa) +
  scale_fill_manual(values = c(Pre = "#4e9af1", Post = "#f17a4e"),
                    name = "Period") +
  theme_bw() +
  labs(title = "Recruitment by habitat and taxa",
       x = "Year", y = "Recruits per 5 m²")

recruit_transect %>%
  filter(taxa != "Mil") %>%
  mutate(site_habitat = paste(site, habitat, sep = "-")) %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = pre_post_f)) +
  geom_col() +
  facet_grid(site_habitat ~ taxa) +
  scale_fill_manual(values = c(Pre = "#4e9af1", Post = "#f17a4e"),
                    name = "Period") +
  theme_bw() +
  labs(title = "Recruitment by site-habitat combination and taxa",
       x = "Year", y = "Recruits per 5 m²")


```

## Model 1: Habitat level recruitment befor and after bleaching

-   **Formula: habitat \* year + (1\|site/transect_id**
-   **site/transect_id: sites pooled as random... island-wide habitat question**
-   **year crossed random effect: absorbs annual variation**


**Get estimate on link scale and then get se since only on link scale and then convert to response space 
** tweedie distribution (keep in decimal format, no transformation (inflated zero), hurdle model too)(has zero inflation built in for hurdle)


### Pocillopera

```{r}
# Poc
cat("Q1: Poc ~ habitat * year | site/transect_id" )
cat("Family: tweedie\n")

# q1_poc <- glmmTMB(
#   recruits_std ~ habitat * pre_post_f + (1 | site/transect_id) + (1 | year),
#   family    = tweedie(link = "log"),
#   data      = dat_poc,
#   na.action = na.exclude
# )

q1_poc_year <- glmmTMB(
  recruits_std ~ habitat * factor(year) + (1 | site/transect_id),
  family    = tweedie(link = "log"),
  data      = dat_poc,
  na.action = na.exclude
)

summary(q1_poc_year) 
```

Interpretation: Pre-bleaching OR recruited more than BR. Post-bleaching had flipped pattern: OR dropped sharply while BR held (interaction is p \< .05). OR recruitment seemed to drop significantly post bleaching.

```{r}
# Prediction grid
poc_pred_grid_year <- dat_poc %>%
  distinct(habitat, year)

poc_pred_grid_year$predicted <- predict(q1_poc_year, newdata = poc_pred_grid_year,type = "response", re.form = NA)

poc_pred_grid_year$predicted

# Need to extract the CI (se)
```

```{r}
# Visualization
# New prediction plot 
# Plot predictions for Poc
ggplot(poc_pred_grid_year, aes(x = factor(year), y = predicted, col = factor(habitat))) + 
  geom_point() + geom_line(group = 1) + facet_wrap(~habitat) + theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  labs(title = "Island-wide Coral Recruitment by Habitat and Bleaching Period Per Taxa")
```

#### Model Validation
```{r}
# Model validation 
## Model validation and assumptions 


# 1. for negative binomial, we will need to test for the overdispersion (qqplot), heteroscedacity, and zero-inflation

# 2. Residual diagnostics (outliers, non-normality, or deviations)
# Plot of predicted vs actual, should not see any curvature, will get noise but on the line mostly 

# Q2A (data never really matches the model... how different is the model from the assumption)
# switch to tweedie model for #Q2a.. try tweedie() 
```


### Porites

```{r}
# Por 
cat("Q1 Por ~ habitat * pre_post_f | site/transect_id + year\n")
cat("Family: nbinom2\n")


q1_por <- glmmTMB(
  recruits_std ~ habitat * pre_post_f + (1 | site/transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_por,
  na.action = na.exclude
)

summary(q1_por)
```

```{r}
por_pred_grid_q1 <- dat_por %>%
  distinct(habitat, pre_post_f)

por_pred_grid_q1$predicted <- predict(q1_por, newdata = por_pred_grid_q1,
                                      type = "response", re.form = NA)

por_pred_grid_q1$predicted
por_se <- predict(q1_por, newdata = por_pred_grid_q1,
                  type = "response", re.form = NA, se.fit = TRUE)
por_pred_grid_q1$se <- por_se$se.fit
```



Opposite pattern seen: BR recruited much more than OR pre-bleaching (habitatOR = -1.91 and is significant). Both habitats dropped post bleaching (pre_post is significant); however, the gap is not significant since interaction is not significant.




```{r}

```




```{r}
# Prediction grids 
# recruits_int ~ habitat * pre_post_f + (1|site/transect_id) + (1|year)
# type = "response"  to back transform from log scale to count scale

# # Poc
# poc_pred_grid_q1 <- dat_poc %>%
#   distinct(habitat, pre_post_f)
# 
# poc_pred_grid_q1$predicted <- predict(q1_poc, newdata = poc_pred_grid_q1,
#                                       type = "response", re.form = NA)
# 
# poc_pred_grid_q1
# 
# 
# # Por
# 
# por_pred_grid_q1 <- dat_por %>%
#   distinct(habitat, pre_post_f)
# 
# por_pred_grid_q1$predicted <- predict(q1_por, newdata = por_pred_grid_q1,
#                                       type = "response", re.form = NA)
# 
# por_pred_grid_q1


# Combine for plotting 

q1_pred_all <- bind_rows(
  acr_pred_grid_q1 %>% mutate(taxa = "Acropora"),
  poc_pred_grid_q1 %>% mutate(taxa = "Pocillopora"),
  por_pred_grid_q1 %>% mutate(taxa = "Porites")
) %>%
  mutate(
    taxa       = factor(taxa, levels = c("Acropora", "Pocillopora", "Porites")),
    pre_post_f = factor(pre_post_f, levels = c("Pre", "Post"))
  )


# Bar plot 
# ******predict() gives point estimates only... so no se (how to get??), use emmeans??
# For se bars need emmeans() or bootstrap 



period_colors <- c(Pre = "#4e9af1", Post = "#f17a4e")

# Combine
q1_pred_all <- bind_rows(
  acr_pred_grid_q1 %>% mutate(taxa = "Acropora"),
  poc_pred_grid_q1 %>% mutate(taxa = "Pocillopora"),
  por_pred_grid_q1 %>% mutate(taxa = "Porites")
) %>%
  mutate(
    taxa       = factor(taxa, levels = c("Acropora", "Pocillopora", "Porites")),
    pre_post_f = factor(pre_post_f, levels = c("Pre", "Post"))
  )




# Plot
ggplot(q1_pred_all, 
       aes(x = habitat, y = predicted, fill = pre_post_f)) +
  geom_col(position = position_dodge(width = 0.75),
           width = 0.65, colour = "white", linewidth = 0.3) +
  facet_wrap(~ taxa, scales = "free_y", ncol = 3) +
  scale_fill_manual(
    values = c(Pre = "#4e9af1", Post = "#f17a4e"),
    name   = "Period",
    labels = c(Pre = "Pre-bleaching (2013-2018)",
               Post = "Post-bleaching (2019-2024)")
  ) +
  labs(
    title = "Island-wide coral recruitment by habitat and bleaching period",
    x     = "Habitat",
    y     = "Predicted recruits per 5 m^2"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

```

Bleaching hit recruitment hard across all three taxa and both habitats. Acr and Poc follow the same habitat preference (OR \> BR pre-bleaching, both crash post-bleaching). Por is the differing one; it's a back reef recruiter that was devastated. Wide CI on Porites BR Pre reflects genuine high variability in that cell.

## Model 2: Island-wide recruitment by site x habitat before and after bleaching 
Do site × habitat differences in recruitment vary between pre and post-bleaching periods?

Formula: site \* habitat \* pre_post_f + (1\|transect_id) + (1\|year)

### Acropora

```{r}
cat("Q2 Acr ~ site * habitat * pre_post_f | transect_id + year\n")

q2a_acr <- glmmTMB(
  recruits_std ~ site * habitat * pre_post_f + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_acr,
  na.action = na.exclude
)

summary(q2a_acr)
hist(dat_acr$recruits_std, breaks = 20)
hist(dat_por$recruits_std, breaks = 20)
hist(dat_poc$recruits_std, breaks = 20)


```

### Pocilopora

```{r}
cat("Q2A Poc ~ site * habitat * pre_post_f | transect_id + year\n")
q2a_poc <- glmmTMB(
  recruits_std ~ site * habitat * pre_post_f + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_poc,
  na.action = na.exclude
)
summary(q2a_poc)

# predictions get standard erorrs, conversion to response space

```

### Porites

```{r}
cat("Q2A Por ~ site * habitat * pre_post_f | transect_id + year\n")
q2a_por <- glmmTMB(
  recruits_std ~ site * habitat * pre_post_f + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_por,
  na.action = na.exclude
)

summary(q2a_por)
```

