---
title: "recruitment_glmm_walkthrough"
format: html
editor_options: 
  chunk_output_type: console
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora from 2013-2024?**

## Data Preparation

### Load and Clean

The following packages were required prior to conducting additional data wrangling:

```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation 
```

Then, the cleaned data (referenced from a separate repository) was loaded in.

```{r}
#| message: false
#| warning: false
# clean_coral <- read_csv( 
# "~/MEDS/capstone/eds-coral-data-storage-management/analysisdata/coral_demography_analysis_data.csv", guess_max = 1000)

clean_coral <- read_csv( 
"~/MEDS/capstone/eds-coral-glmms/recruitment/data_recruitment/tableau_recruitment_coral_demography_analysis_data.csv", guess_max = 1000 )

## Log size
tableau_data <- clean_coral %>%
  mutate(size_viz_log = ifelse(!is.na(size_viz), log10(size_viz + 1), NA))

# write_csv(tableau_data, "~/MEDS/capstone/eds-coral-glmms/recruitment/data_recruitment/tableau_clean.csv")

tableau_data <- clean_coral %>%
  group_by(coral_id) %>%
  arrange(year) %>%
  mutate(
    ever_died_int = as.integer(cumsum(replace_na(dyn_death, 0)) > 0),
    size_viz = case_when(
      ever_died_int == 1 ~ NA_real_,        # dead → disappear
      !is.na(size_t1) ~ size_t1,            # alive with size_t1
      !is.na(size_t0) ~ size_t0,            # alive with size_t0
      TRUE ~ size_t1                              # alive but no measurement → small dot
    )
  ) %>%
  ungroup()

# Verify
tableau_data %>%
  filter(coral_id == "LTER1-BR-T01-Acr-0.02-55-65-2013") %>%
  select(coral_id, year, ever_died_int, size_viz)

# Export
write_csv(tableau_data, "~/MEDS/capstone/eds-coral-glmms/recruitment/data_recruitment/tableau_clean.csv")

# Take a look at the cleaned coral dataset
glimpse(clean_coral)
unique(clean_coral$transect_id)

view(clean_coral)
```

```{r}
# Rename habitats 
clean_coral$habitat <- fct_recode(clean_coral$habitat, 
                "Back Reef" = "BR", 
                "Fore Reef" = "OR")

# Integrate extra transects
clean_coral <- clean_coral %>% mutate(transect_id = case_when(
      transect_id == "LTER1_OR_TO1"         ~ "LTER1_OR_T01",
      transect_id == "LTER1_OR_T02_2021"    ~ "LTER1_OR_T02",
      transect_id == "LTER2_OR_T04_OLD"     ~ "LTER2_OR_T04",
      transect_id == "LTER5_OR_T02_OLD"     ~ "LTER5_OR_T02",
      transect_id == "LTER5_OR_T03_PRE2020" ~ "LTER5_OR_T03",
      TRUE ~ transect_id
     ))
unique(clean_coral$transect_id)
view(clean_coral)

```

Check the bounds of Poc and Por in the fore reef.

```{r}
# Poc forereef
# view(clean_coral %>%
#   filter(habitat == "Fore Reef", taxa == "Poc", dyn_recruitment == 1) %>%
#   group_by(site, year, transect) %>%
#   summarise(
#     max_x = max(x, na.rm = TRUE),
#     n_beyond_1m = sum(x > 1, na.rm = TRUE),
#     total = n(),
#     .groups = "drop"
#   ) %>%
#   filter(n_beyond_1m > 0) %>%        # only show transects that actually have points outside 1m
#   arrange(site, year, transect) %>%
#   print(n = Inf))
# 
# view(clean_coral %>% filter(habitat == "Fore Reef", taxa == "Poc", site == "LTER1") %>% select(dyn_recruitment))
# # Por forereef
# clean_coral %>%
#   filter(habitat == "Fore Reef", taxa == "Por", dyn_recruitment == 1) %>%
#   group_by(site, year) %>%
#   summarise(max_x = max(x, na.rm = TRUE),
#             n_beyond_1m = sum(x > 1, na.rm = TRUE),
#             total = n(), .groups = "drop") %>%
#   arrange(site, year) %>%
#   print(n = Inf)
```

```{r}
# Are there any NON recruits (corals already there)?
clean_coral %>%
  filter(taxa == "Poc", habitat == "Fore Reef") %>%
  group_by(dyn_recruitment) %>%
  summarize(
    n = n(),
    pct = round(100 * n() / nrow(clean_coral %>% 
                filter(taxa == "Poc", habitat == "Fore Reef")), 1)
  )
```

```{r}
#Data checks
# 1. Total observations before and after filtering
# cat("Total rows before filter:", nrow(clean_coral), "\n")
# cat("Total rows after filter:", nrow(clean_filtered), "\n")
# cat("Rows removed:", nrow(clean_coral) - nrow(clean_filtered), "\n\n")
# 
# # 2. Breakdown of removed rows
# clean_coral %>%
#   filter(taxa == "Poc", habitat == "Fore Reef", x > 1) %>%
#   summarize(
#     total_removed      = n(),
#     recruits_removed   = sum(dyn_recruitment == 1),
#     nonrecruits_removed = sum(dyn_recruitment == 0)
#   ) %>% print()
# 
# # Por recruits removed
# clean_coral %>%
#   filter(taxa == "Por", habitat == "Fore Reef", x > 1) %>%
#   summarize(
#     total_removed      = n(),
#     recruits_removed   = sum(dyn_recruitment == 1),
#     nonrecruits_removed = sum(dyn_recruitment == 0)
#   ) %>% print()
# 
# 
# 
# # 3. OR Poc recruit retention rate
# cat("\nOR Poc recruit retention:\n")
# cat("Before filter:", 901, "recruits\n")
# cat("After filter:", 901 - 26, "recruits\n")
# cat("Retention rate:", round(100 * (901 - 26) / 901, 1), "%\n\n")
# 
# # 4. Transects affected
# cat("Transects losing some recruits after filter:\n")
# before <- clean_coral %>%
#   filter(taxa == "Poc", habitat == "Fore Reef", dyn_recruitment == 1) %>%
#   group_by(transect_id) %>%
#   summarize(n_before = n(), .groups = "drop")
# 
# after <- clean_filtered %>%
#   filter(taxa == "Poc", habitat == "Fore Reef", dyn_recruitment == 1) %>%
#   group_by(transect_id) %>%
#   summarize(n_after = n(), .groups = "drop")
# 
# left_join(before, after, by = "transect_id") %>%
#   mutate(lost = n_before - replace_na(n_after, 0)) %>%
#   filter(lost > 0) %>%
#   print(n = Inf)
```

Pocillopora recruits on the Fore Reef were sampled within the first 1m² of each 5m² transect. A total of 26 recruit observations (2.9% of Fore Reef Pocillopora recruits) recorded beyond the 1m² sampling boundary were identified as data entry errors and excluded from analysis following confirmation from the field data collector.

### Recruitment Standardization

Standardize recruitment for every meters squared on the POC on the forereef

```{r}
# Exclude the Poc recruits OUTSIDE 1 m^2
recruit_transect <- clean_coral %>%
 # mutate(x = ifelse(x > 5, x / 10, x)) %>%
  filter(!(habitat == "Fore Reef" & taxa == c("Poc", "Por") & x > 1)) %>% 
           #& dyn_recruitment == 1)) %>%
  group_by(taxa, site, habitat, transect_id, year) %>%
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    n_recruits = ifelse(year == 2017, NA, n_recruits),
    recruits_std = case_when(
      as.character(habitat) == "Fore Reef" & as.character(taxa) %in% c("Poc", "Por") ~ n_recruits * 5,
      TRUE ~ n_recruits
    ),
    bleaching_status = factor(
      case_when(
        year >= 2013 & year <= 2018 ~ "Pre",
        year >= 2019 & year <= 2024 ~ "Post"
      ),
      levels = c("Pre", "Post")
    ),
    taxa    = factor(taxa),
    site    = relevel(factor(site), ref = "LTER1"),
    habitat = relevel(factor(habitat), ref = "Back Reef")
  ) 
#%>%filter(!is.na(bleaching_status))

# Quick verify
glimpse(recruit_transect)
unique(recruit_transect$habitat)

# Any out of bounds corals? 
# clean_filtered %>%
#   mutate(x = ifelse(x > 5, x / 10, x)) %>%
#   filter(x > 5) %>%
#   count(taxa, habitat, site)


# Check recruitment 
#view(recruit_transect)

write.csv(recruit_transect, file = "recruitment/data_recruitment/recruits_transect.csv")


```

\*\*\*\*Offshore for poc are recruits are counted in the first m\^2 (check and make sure this is the case, so not find recruits past the first meter sites)

```{r}
# Create filtered datasets for each taxa
dat_acr <- recruit_transect %>% filter(taxa == "Acr")
dat_poc <- recruit_transect %>% filter(taxa == "Poc")
dat_por <- recruit_transect %>% filter(taxa == "Por")
```

```{r}
# Exploratory bar charts 
# recruit_transect %>%
#   filter(taxa != "Mil") %>%
#   ggplot(aes(x = factor(year), y = , fill = bleaching_status)) +
#   geom_col() +
#   facet_grid(habitat ~ taxa, scales = "free_y") +  # Not constraint by y-axis
#   scale_fill_manual(values = c(Pre = "dodgerblue", Post = "red"),
#                     name = "Period") +
#   scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
#   theme_bw() +
#   labs(title = "Recruitment by Habitat and Taxa",
#        x = "Year", y = "Recruits per 5 m²") + 
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# recruit_transect %>%
#   filter(taxa != "Mil") %>%
#   mutate(site_habitat = paste(site, habitat, sep = "-")) %>%
#   ggplot(aes(x = factor(year), y = recruits_std, fill = bleaching_status)) +
#   geom_col() +
#   facet_grid(site_habitat ~ taxa) +
#   scale_fill_manual(values = c(Pre = "dodgerblue", Post = "red"),
#                     name = "Period") +
#   theme_bw() +
#   labs(title = "Recruitment by site-habitat combination and taxa",
#        x = "Year", y = "Recruits per 5 m²") + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


recruit_transect %>%
    #filter(taxa != "Mil") %>%
    filter(!is.na(recruits_std)) %>%
    ggplot(aes(x = taxa, y = recruits_std, fill = taxa)) +
    geom_boxplot(outlier.shape = 21, outlier.size = 1.5) +
    scale_y_log10() +  # Log scale since Por BR 2013 will dominate
    scale_fill_manual(values = c(Acr = "#E63946", Poc = "#F4A6C0", Por = "#D4A017", Mil = "#FF7F2A")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(
        title   = "Distribution of Coral Recruitment by Taxa",
        x       = "Taxa",
        y       = "Recruits per 5 m² (log scale)",
        caption = "*Each observation = one transect-year (2017 excluded)"
    )

#****Each observation = one transect-year. 2017 excluded
```

```{r}
# Acr
dat_acr %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  + 
    labs(title = "Habitat-wide Coral Recruitment of Acropora (2013-2024)", caption = "2017 not sampled.") + theme(legend.position = "none")


dat_por %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
    labs(title = "Habitat-wide Coral Recruitment of Porites (2013-2024)", caption = "2017 not sampled.") + theme(legend.position = "none")


dat_poc %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  +
    labs(title = "Habitat-wide Coral Recruitment of Pocillopera (2013-2024)", caption = "2017 not sampled.") 



```



## Model 1: Habitat level recruitment through time

How does coral recruitment differ between lagoon and forereef habitats across years 2013-2024?

-   **Formula: habitat \* year + (1\|site/transect_id)**
-   **site/transect_id: sites pooled as random... island-wide habitat question**
-   **year crossed random effect: absorbs annual variation**


**SITE IS NOW A FIXED EFFECT TO CONTROL FOR GROUPING ALL TRANSECTS AND SITES TOGETHER (KEEP THEM DISTINCT!!)

\*\*Get estimate on link scale and then get se since only on link scale and then convert to response space

How did recruitment differ in the lagoon and forereef between 2013 and 2024? Recruitment... most abundant genera of corals (Por, Poc, Acr)... How you did it? (Poc is for 1 m squared of forereef)

### Acropora

Year-to-year for LONG term declines. ZiG: conditional is if there are no zeroes (ie. it was not there) and zero-inflated is with those zeroes (flexible). In order to interpret, will need to understand the predictions?

```{r}
#| collapse: true

q1_acr_year <- glmmTMB(
  recruits_std ~ habitat*year + site + (1 | transect_id),
  #family = tweedie(link = "log"),
  family = ziGamma(link = "log"),
   ziformula = ~.,
  data      = dat_acr,
  na.action = na.exclude
)

summary(q1_acr_year)

q1_acr_bleaching_status <- glmmTMB(
  recruits_std ~ habitat*bleaching_status + (1 | site/transect_id) + (1|year),
  #family = tweedie(link = "log"),
  family = ziGamma(link = "log"),
  ziformula = ~.,
  data      = dat_acr,
  na.action = na.exclude
)

summary(q1_acr_bleaching_status)

q1_poc_bleaching_status <- glmmTMB(
  recruits_std ~ habitat*bleaching_status + (1 | site/transect_id) + (1|year),
  family = tweedie(link = "log"),
 # family = ziGamma(link = "log"),
  # ziformula = ~.,
  data      = dat_poc,
  na.action = na.exclude
)

summary(q1_poc_bleaching_status)


q1_por_bleaching_status <- glmmTMB(
  recruits_std ~ habitat*bleaching_status + (1 | site/transect_id) + (1|year),
  family = tweedie(link = "log"),
 # family = ziGamma(link = "log"),
  # ziformula = ~.,
  data      = dat_por,
  na.action = na.exclude
)
summary(q1_por_bleaching_status)


#view(dat_acr)
# Visualize predictions
# Summarize standardized recruits for each habitat and year
acr_avg_recruits_year <- dat_acr %>% group_by(habitat,year) %>%
  summarize(
    #total_recruits = sum(recruits_std, na.rm = T)) %>% ungroup()
    avg_recruits = mean(recruits_std, na.rm = T)) %>% ungroup()

acr_prediction_grid <- dat_acr %>%
  filter(year != 2017) %>%
  distinct(habitat, year)

acr_prediction_grid$predicted<- predict(q1_acr_year,
                                        newdata = acr_prediction_grid,
                                        type = "response", re.form = NA)

# Join predictions to observed
acr_plot_data <- acr_avg_recruits_year %>% left_join(acr_prediction_grid, by = c("habitat", "year"))
view(acr_plot_data)
# Plot
ggplot(acr_plot_data) +
  geom_col(aes(x = factor(year), y = avg_recruits, fill = habitat),
           alpha = 0.5) +
  geom_point(aes(x = factor(year), y = predicted, col = habitat)) +
  geom_line(aes(x = factor(year), y = predicted, col = habitat,
                group = habitat)) +
  facet_wrap(~ habitat) +
  scale_fill_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),
    name   = "Observed"
  ) +
  scale_color_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),
    name   = "Predicted"
  ) +
  scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
  theme_bw() +
  theme( # Specify mean observed and predicted recruits
    axis.text.x   = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    title   = "Habitat-wide Coral Recruitment of Acropora (2013-2024)",
    y       = "Recruits per 5 m²",
    x       = "Year",
    caption = "Bars = observed mean, 2017 not sampled."
  )
```

```{r}
# Add error bars
acr_summary <- dat_acr %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, year, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(acr_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Acropora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 


  
  
```

```{r}
## habitat-bleaching
acr_summary <- dat_acr %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(acr_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Acropora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 
```




### Pocillopera

```{r}
# Poc
q1_poc_year <- glmmTMB(
  recruits_std ~ habitat * factor(year) + site + (1 | transect_id),
  family    = tweedie(link = "log"),
  data      = dat_poc,
  na.action = na.exclude
)


summary(q1_poc_year) 

```

```{r}
## habitat-bleaching
poc_summary <- dat_poc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(poc_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    # scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Period",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Pocillopora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 
```



```{r}
# Add error bars
poc_summary <- dat_poc %>%
    #filter(!is.na(recruits_std)) %>%
    group_by(habitat, year, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(poc_summary,
       #%>% 
         #filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec ,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Pocillopora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) + theme(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA),
  legend.background = element_rect(fill = "transparent", colour = NA),
  legend.box.background = element_rect(fill = "transparent", colour = NA)
)
```



```{r}
# Add error bars
poc_summary <- dat_poc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, year, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(poc_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec - se_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m² (+SE)",
        title = "Habitat-wide Coral Recruitment of Pocillopora (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    )
```




```{r}
#| message: false
#| warning: false
# Visualization
# Observed data (keep all years including 2017 and get one mean observation for habitat-year)
poc_avg_recruits_year <- dat_poc %>% 
  group_by(habitat, year) %>% 
  summarize(avg_recruit = mean(recruits_std, na.rm = TRUE)) %>% 
  ungroup() 

# Predictions (2017 excluded) 
poc_pred_grid_year <- dat_poc %>%
  filter(year != 2017) %>%
  distinct(habitat, year)

poc_pred_grid_year$predicted <- predict(q1_poc_year, 
                                        newdata = poc_pred_grid_year,
                                        type = "response",
                                        re.form = NA)

# Join observed to predictions
poc_plot_data <- poc_avg_recruits_year %>%
  left_join(poc_pred_grid_year, by = c("habitat", "year"))

# Plot
ggplot(poc_avg_recruits_year) + 
  geom_col(aes(x = factor(year), y = avg_recruit, fill = habitat), 
           alpha = 0.5) +
  # geom_point(aes(x = factor(year), y = predicted, col = habitat)) + 
  # geom_line(aes(x = factor(year), y = predicted, col = habitat, 
  #               group = habitat)) +  
  facet_wrap(~ habitat) +
  scale_fill_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),  
    name   = "Observed"
  ) +
  # scale_color_manual(
  #   values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"), 
  #   name   = "Predicted"
  # ) +
  scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
  theme_bw() + 
  theme( # Specify mean observed and predicted recruits
    axis.text.x   = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    title   = "Coral Recruitment of Pocillopora Between Habitats (2013-2024)",
    y       = "Recruits per 5 m²",
    x       = "Year",
    caption = "Bars = observed mean. 2017 not sampled."
  )
```

Pocillopora recruitment differed significantly between the back reef and fore reef across the study period, with the fore reef recruiting significantly more overall (p \< 0.001). Back reef recruitment remained relatively stable between 2013 and 2019, with only 2018 exhibiting a notable increase (p = 0.035). A significant decline in recruitment followed by 2021 (p = 0.004).

The fore reef showed a striking decline in recruitment over the time period. The fore reef was the dominant recruitment habitat in 2013 but declined sharply relatively to the back reef from 2014 onwards (p = 0.028). This divergence in recruitment away from the back reef became increasingly significant through the post-bleaching year; every year from 2018-2024 showed a highly significant reduction in fore reef recruitment relative to 2013 (all p \< 0.001) except 2016 (p = 0.016). Recruitment was worse in the fore reef in 2022 (p \< 0.001) and did not show signs of significant recovery through 2024 (p \< 0.001).

#### Model Validation

```{r}
# Model validation 
## Model validation and assumptions 


# 1. for negative binomial, we will need to test for the overdispersion (qqplot), heteroscedacity, and zero-inflation

# 2. Residual diagnostics (outliers, non-normality, or deviations)
# Plot of predicted vs actual, should not see any curvature, will get noise but on the line mostly 

# Q2A (data never really matches the model... how different is the model from the assumption)
# switch to tweedie model for #Q2a.. try tweedie() 
```

### Porites

```{r}
# Por 
q1_por_year <- glmmTMB(
  recruits_std ~ habitat * factor(year) + site + (1 | transect_id),
  family    = tweedie(link = "log"),
  data      = dat_por,
  na.action = na.exclude
)

summary(q1_por_year)

```

```{r}
por_summary <- dat_por %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, year, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(por_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Porites Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    )
```


```{r}
# Visualization
# Observed data (keep all years including 2017)
por_avg_recruits_year <- dat_por %>% 
  group_by(habitat, year) %>% 
  summarize(avg_recruit = mean(recruits_std, na.rm = TRUE)) %>% 
  ungroup()

# Predictions (2017 excluded) 
por_pred_grid_year <- dat_por %>%
  filter(year != 2017) %>%
  distinct(habitat, year)

por_pred_grid_year$predicted <- predict(q1_por_year, 
                                        newdata = por_pred_grid_year,
                                        type = "response",
                                        re.form = NA)

# Join observed to predictions
por_plot_data <- por_avg_recruits_year %>%
  left_join(por_pred_grid_year, by = c("habitat", "year"))

# Plot
ggplot(por_plot_data) + 
  geom_col(aes(x = factor(year), y = avg_recruit, fill = habitat), 
           alpha = 0.5) +
  geom_point(aes(x = factor(year), y = predicted, col = habitat)) + 
  geom_line(aes(x = factor(year), y = predicted, col = habitat, 
                group = habitat)) +  
  facet_wrap(~ habitat) +
  scale_fill_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),  
    name   = "Observed"
  ) +
  scale_color_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"), 
    name   = "Predicted"
  ) +
  scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
  theme_bw() + 
  theme( # Specify mean observed and predicted recruits
    axis.text.x   = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    title   = "Habitat-wide Coral Recruitment of Porites (2013-2024)",
    y       = "Recruits per 5 m²",
    x       = "Year",
    caption = "Bars = observed mean, points = model predicted. 2017 not sampled."
  )
```


```{r}
## habitat-bleaching
por_summary <- dat_por %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(por_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    # scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Period",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Porites Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 
```




Porites recruitment was significantly higher in the back reef than the forereef across the entire study period (p \< 0.001), making it the only taxa to show back reef dominance. Back reef recruitment was exceptionally high in 2013 then declined sharply and significantly every subsequent year (all p \< 0.001), with the steepest drops occurring after 2019 and bottoming out in 2024, worst year on record (p \< 0.001). This monotonic decline suggests a long-term suppression of back reef Porites recruitment that predates the bleaching event and worsened dramatically after it.

The forereef told a different story. While it recruited significantly less than the back reef overall, the gap between habitats narrowed significantly in 2018 and 2019 (p = 0.021 and p = 0.037) as the back reef collapsed faster than the forereef during the bleaching period. By 2024 the interaction term reversed significantly (p = 0.009), not because the forereef recovered, but because the back reef had declined so severely that the two habitats converged near zero. This convergence reflects a system-wide collapse of Porites recruitment rather than any habitat-specific recovery, consistent with the near-zero values visible across both habitats in the post-2019 period.Model Validation

## Model 2: Island-wide recruitment by site x habitat before and after bleaching

Do site × habitat differences in recruitment vary between pre and post-bleaching periods?

Formula: site \* habitat \* bleaching_status + (1\|transect_id) + (1\|year)

### Acropora
Transect_id is dropped here because every time transect_id was included it collapsed to essentially zero (1.367e-11, 1.431e-09)... means data couldn't distinguish between transects for Acropora. Sparse counts with many zeros meant there wasn't enough within-transect variation to estimate a transect-level random effect. Dropping it gave a stable and cleaner model.

```{r}
q2a_acr <- glmmTMB(
  recruits_std ~ site*habitat*bleaching_status +  (1 | year),
  # family = tweedie(link = "log"),
 family    = ziGamma(link = "log"),
 ziformula = ~habitat + bleaching_status, # this combo has more var in recruits
  data      = dat_acr,
  na.action = na.exclude
)

summary(q2a_acr)

# ) %>%
# LTER 4 FOR REEF barely has zeros... variance is so low
# A tibble: 16 × 6
#    site  habitat   bleaching_status n_total n_zeros pct_zeros
#    <fct> <fct>     <fct>              <int>   <int>     <dbl>
#  1 LTER1 Back Reef Pre                   20      14      70  
#  2 LTER1 Back Reef Post                  24      13      54.2
#  3 LTER1 Fore Reef Pre                   20       3      15  
#  4 LTER1 Fore Reef Post                  24      16      66.7
#  5 LTER2 Back Reef Pre                   20      14      70  
#  6 LTER2 Back Reef Post                  24      18      75  
#  7 LTER2 Fore Reef Pre                   20       4      20  
#  8 LTER2 Fore Reef Post                  24      10      41.7
#  9 LTER4 Back Reef Pre                   20       9      45  
# 10 LTER4 Back Reef Post                  24       8      33.3
# 11 LTER4 Fore Reef Pre                   20       0       0  
# 12 LTER4 Fore Reef Post                  24       7      29.2
# 13 LTER5 Back Reef Pre                   20      18      90  
# 14 LTER5 Back Reef Post                  24      20      83.3
# 15 LTER5 Fore Reef Pre                   20       5      25  
# 16 LTER5 Fore Reef Post                  24       6      25  


```

```{r}
# Prediction grid
acr_pred_grid_q2 <- dat_acr %>%
  distinct(site, habitat, bleaching_status)

acr_fit_q2 <- predict(q2a_acr, newdata = acr_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

acr_pred_grid_q2$predicted <- acr_fit_q2$fit
acr_pred_grid_q2$se <- acr_fit_q2$se.fit

# Observed means 
acr_obs_q2 <- dat_acr %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")

view(acr_obs_q2)
view(acr_pred_grid_q2)

```



```{r}
acr_summary_q2 <- dat_acr %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()), # se
    .groups = "drop"
  )


ggplot(acr_summary_q2,
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec,
                    ymax = mean_rec + se_rec),
                width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    labels = c(Pre = "Pre-bleaching", Post = "Post-bleaching")
  ) + 
  theme_bw() +
  labs(
    title = "Acropora Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Bleaching Period",
    y = "Mean recruits per 5 m²",
    fill = "Bleaching Period"
  ) + scale_x_discrete(labels = c(Pre = "Pre-bleaching\n(2013-2018)", 
                             Post = "Post-bleaching\n(2019-2024)")) +   theme(axis.title.x = element_blank())

```



### Pocillopora

```{r}
q2a_poc <- glmmTMB(
  recruits_std ~ site * habitat * bleaching_status + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_poc,
  na.action = na.exclude
)
summary(q2a_poc)

# predictions to get standard errors, conversion to response space!!! 

```

```{r}
# Calculate the average observed recruits per transect at site-habitat-period level
poc_summary_q2 <- dat_poc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()), # se
    .groups = "drop"
  )


ggplot(poc_summary_q2,
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
  geom_col(width = 0.6, colour = "white") +
  geom_errorbar(aes(ymin = mean_rec,
                    ymax = mean_rec + se_rec),
                width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    labels = c(Pre = "Pre-bleaching", Post = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Pocillopora Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) + scale_x_discrete(labels = c(Pre = "Pre-bleaching\n(2013-2018)", 
                             Post = "Post-bleaching\n(2019-2024)")) +   theme(axis.text.x  = element_blank())

```



```{r}
# Prediction grid
poc_pred_grid_q2 <- dat_poc %>%
  distinct(site, habitat, bleaching_status)

poc_fit_q2 <- predict(q2a_poc, newdata = poc_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

poc_pred_grid_q2$predicted <- poc_fit_q2$fit
poc_pred_grid_q2$se <- poc_fit_q2$se.fit

# Observed means 
poc_obs_q2 <- dat_poc %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")
view(poc_obs_q2)
view(poc_pred_grid_q2)

# Combine 
# poc_q2_plot_df <- poc_pred_grid_q2 %>%
#   left_join(poc_obs_q2, by = c("site", "habitat", "bleaching_status")) %>%
#   pivot_longer(cols      = c(predicted, observed),
#                names_to  = "type",
#                values_to = "recruits") %>%
#   mutate(
#     bleaching_status = factor(bleaching_status, levels = c("Pre", "Post")),
#     type       = factor(type, levels = c("observed", "predicted")),
#     site       = factor(site, levels = c("LTER1", "LTER2", "LTER4", "LTER5"))
#   )


```

```{r}
# Predictions
ggplot() +
  # Observed bars
  geom_col(data = poc_obs_q2,
           aes(x = habitat, y = observed, fill = bleaching_status),
           position = position_dodge(width = .5), width = .5, col = "white", linewidth = 0.3, alpha = .7) +
  # # Predicted points
  # geom_point(data = poc_pred_grid_q2,
  #            aes(x = habitat, y = predicted, group = bleaching_status),
  #            position = position_dodge(width = .5),
  #            size = 3, shape = 18) +  
  # Predicted ci
  # geom_errorbar(
  #   data = poc_pred_grid_q2,
  #   aes(x    = habitat,
  #       ymin = exp(log(predicted) - 1.96 * (se / predicted)),
  #       ymax = exp(log(predicted) + 1.96 * (se / predicted)),
  #       group = bleaching_status),
  #   position = position_dodge(width = 0.75),
  #   width = 0.25, linewidth = 0.6
  # ) +
  facet_wrap(~ site) +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    name   = "Period",
    labels = c(Pre = "Pre-bleaching (2013-2018)",
               Post = "Post-bleaching (2019-2024)")
  ) +
  labs(
    title    = "Pocillopora Island-Wide Recruitment",
    caption = "Bars = observed mean, point = model predicted",
    x        = "Habitat",
    y        = "Recruits per 5 m²"
  ) +
  theme_bw() +
  theme(
    legend.position    = "bottom"
  )



```

Prior to bleaching, Pocillopora recruitment was significantly higher on the fore reef than the back reef island-wide (p = .008). There were no significant differences among sites within either habitat, suggesting that there is a consistent habitat-level pattern across Mo'orea. Back reef recruitment remained low and stable across all four sites pre-bleaching while fore reef recruitment was higher at all sites relative to their respective back reefs.

Recruitment patterns in Pocillopora shifted significantly in both habitats. Fore reef recruitment declined sharply at both sites, and the previous advantage for recruitment held in the fore reef disappeared at LTER1 (p \< 0.001). Recruitment also declined significantly at the LTER 2 back reef (p = 0.049). However, the degree of this decline varied among sites. LTER 2 fore reef experienced a smaller decline than LTER1 (p = 0.031). LTER 5 fore reef showed no significant decline post-bleaching, maintaining pre-bleaching recruitment levels (p \< 0.001). Overall, bleaching reduced fore reef recruitment across the island, but some sites (especially LTER 5) appeared to serve as a refuge that could potentially help Pocillopora recovery.

### Porites

```{r}
q2a_por <- glmmTMB(
  recruits_std ~ site * habitat * bleaching_status + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_por,
  na.action = na.exclude
)
summary(q2a_por)

```

```{r}
# Prediction grid
por_pred_grid_q2 <- dat_por %>%
  distinct(site, habitat, bleaching_status)

por_fit_q2 <- predict(q2a_por, newdata = por_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

por_pred_grid_q2$predicted <- por_fit_q2$fit
por_pred_grid_q2$se <- por_fit_q2$se.fit

# Observed means 
por_obs_q2 <- dat_por %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")
view(por_obs_q2)
view(por_pred_grid_q2)
```


```{r}
# Calculate the average observed recruits per transect at site-habitat-period level
por_summary_q2 <- dat_por %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()), # se
    .groups = "drop"
  )


ggplot(por_summary_q2,
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec,
                    ymax = mean_rec + se_rec),
                width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    labels = c(Pre = "Pre-bleaching", Post = "Post-bleaching")
  ) +
  theme_bw() + facet_grid(habitat~site) + 
  labs(
    title = "Porites Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Bleaching Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) + scale_x_discrete(labels = c(Pre = "Pre-bleaching\n(2013-2018)", 
                             Post = "Post-bleaching\n(2019-2024)"))
```


Porites recruitment showed an opposing pattern to Pocillopora; there was notably much higher recruitment in the back reef compared to the fore reef pre-bleaching. This pattern was clear at LTER1 (p = 0.019), consistent with this taxa’s preference for lagoonal back reefs. The preference for back reef habitat was stronger at LTER4 and LTER5, where fore reef recruitment was much lower than at LTER1 fore reef (p \< 0.001 and p = 0.003). Post-bleaching, back reef recruitment declined significantly at LTER1 (p \< 0.001), LTER2 (p = 0.003), and LTER5 (p = 0.026). This significant decrease indicates a widespread decline in Porites’ main recruitment habitat.

Fore reef recruitment was already low before bleaching and declined further at most sites post-bleaching. However, LTER5 appeared to be relatively unchanged in recruitment at the fore reef. A significant interaction between habitat and bleaching indicated that fore reef recruitment declined more strongly than back reef recruitment at LTER1 (p = 0.012). Site-level differences showed importance: LTER2 fore reef was less affected by bleaching compared to LTER1 (p \< 0.024), while recruitment remained similar before and after bleaching at LTER5 fore reef (p \< 0.001). LTER4 showed a similar but non-significant trend (p = 0.067). At LTER5, back reef Porites recruitment declined significantly post-bleaching (p = 0.026), as contrasted with fore reef recruitment at this site.

