---
title: "recruitment_glmm_walkthrough"
format: html
editor_options: 
  chunk_output_type: console
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora from 2013-2024?**

## Data Preparation

### Load and Clean

The following packages were required prior to conducting additional data wrangling:

```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation
library(emmeans) # For estimating predicted recruits 
```

Then, the cleaned data (referenced from a separate repository) was loaded in.

```{r}
#| message: false
#| warning: false
#| 
clean_coral <- read_csv( 
"~/MEDS/capstone/eds-coral-glmms/recruitment/data_recruitment/tableau_recruitment_coral_demography_analysis_data.csv", guess_max = 1000 )

tableau_data <- clean_coral %>%
  group_by(coral_id) %>%
  arrange(year) %>%
  mutate(
    ever_died_int = as.integer(cumsum(replace_na(dyn_death, 0)) > 0),
    size_viz = case_when(
      ever_died_int == 1 ~ NA_real_,        # dead → disappear
      !is.na(size_t1) ~ size_t1,            # alive with size_t1
      !is.na(size_t0) ~ size_t0,            # alive with size_t0
      TRUE ~ size_t1                              # alive but no measurement → small dot
    )
  ) %>%
  ungroup()

# Verify
tableau_data %>%
  filter(coral_id == "LTER1-BR-T01-Acr-0.02-55-65-2013") %>%
  select(coral_id, year, ever_died_int, size_viz)

# Export
write_csv(tableau_data, "~/MEDS/capstone/eds-coral-glmms/recruitment/data_recruitment/tableau_clean.csv")

# Take a look at the cleaned coral dataset
glimpse(clean_coral)
unique(clean_coral$transect_id)

view(clean_coral)
```

```{r}
# Rename habitats 
clean_coral$habitat <- fct_recode(clean_coral$habitat, 
                "Lagoon" = "BR", 
                "Forereef" = "OR")

# Integrate extra transects
clean_coral <- clean_coral %>% mutate(transect_id = case_when(
      transect_id == "LTER1_OR_TO1"         ~ "LTER1_OR_T01",
      transect_id == "LTER1_OR_T02_2021"    ~ "LTER1_OR_T02",
      transect_id == "LTER2_OR_T04_OLD"     ~ "LTER2_OR_T04",
      transect_id == "LTER5_OR_T02_OLD"     ~ "LTER5_OR_T02",
      transect_id == "LTER5_OR_T03_PRE2020" ~ "LTER5_OR_T03",
      TRUE ~ transect_id
     ))
unique(clean_coral$transect_id)
view(clean_coral)

```

Check the bounds of Poc and Por in the fore reef.

```{r}
# Are there any NON recruits (corals already there)?
clean_coral %>%
  filter(taxa %in% c("Poc", "Por"), habitat == "Fore Reef") %>%
  group_by(taxa, dyn_recruitment) %>%
  summarize(
    n = n(),
    pct = round(100 * n() / nrow(clean_coral %>% 
                filter(taxa %in% c("Poc", "Por"), habitat == "Fore Reef")), 1)
  )
```

```{r}
# Data checks
poc_before <- clean_coral %>%
  filter(taxa == "Poc", habitat == "Forereef", dyn_recruitment == 1) %>%
  nrow()

poc_after <- clean_coral %>%
  filter(taxa == "Poc", habitat == "Forereef", dyn_recruitment == 1, x <= 1) %>%
  nrow()

cat("Poc Forereef recruits before filter:", poc_before, "\n")
cat("Poc Forereef recruits after filter:", poc_after, "\n")
cat("Recruits removed:", poc_before - poc_after, "\n")
cat("Retention rate:", round(100 * poc_after / poc_before, 1), "%\n")
```

```{r}
# How many Por observations were excluded by the x > 1 filter
clean_coral %>%
  filter(taxa == "Por", habitat == "Forereef", x > 1) %>%
  summarize(
    total_removed       = n(),
    recruits_removed    = sum(dyn_recruitment == 1, na.rm = TRUE),
    nonrecruits_removed = sum(dyn_recruitment == 0, na.rm = TRUE)
  )

# Retention rate for Por Fore Reef recruits
por_before <- clean_coral %>%
  filter(taxa == "Por", habitat == "Forereef", dyn_recruitment == 1) %>%
  nrow()

por_after <- clean_coral %>%
  filter(taxa == "Por", habitat == "Forereef", dyn_recruitment == 1, x <= 1) %>%
  nrow()

cat("Por Forereef recruits before filter:", por_before, "\n")
cat("Por Forereef recruits after filter:", por_after, "\n")
cat("Recruits removed:", por_before - por_after, "\n")
cat("Retention rate:", round(100 * por_after / por_before, 1), "%\n")
```

*Pocillopora* and *Porites* recruits on the forereef were sampled within the first 1 m² of each 5 m² transect, according to the sampling protocol. A total of 26 *Pocillopora* recruit observations (2.9% of forereef *Pocillopora* recruits) and 24 *Porites* recruit observations (7.8% of forereef *Porites* recruits) recorded beyond the 1 m² sampling boundary were excluded from analysis as they fell outside the defined sampling unit. These exclusions were confirmed with the field data collector prior to analysis.

### Recruitment Standardization

Standardizing was critical to ensure that there are an apples-to-apples comparison

Standardize recruitment for every meters squared on the POC on the forereef

```{r}
# Create standardized recruits variables
recruit_transect <- clean_coral %>%
  filter( # Exclude Poc and Por recruits outside 1 m^2 in forereef
    !(habitat == "Forereef" 
      & taxa %in% c("Poc", "Por") 
      & x > 1 & dyn_recruitment == 1), 
    taxa != "Mil") %>% # Exclude Milopora 
  group_by(taxa, site, habitat, transect_id, year,period) %>% 
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    n_recruits = ifelse(year == 2017, NA, n_recruits),
  recruits_std = case_when(
  habitat == "Forereef" & taxa %in% c("Poc", "Por") ~ n_recruits,        # already 1m²
  TRUE ~ n_recruits / 5) ,                                                  # everything else is 5m²  # Forereef Poc/Por already in 1m²
  #     recruits_std = case_when(
  # habitat == "Forereef" & taxa %in% c("Poc", "Por") ~ n_recruits * 5,
  #      TRUE ~ n_recruits) 
    ,
    bleaching_status = factor(
      case_when(
        period == 0 ~ "Pre",
        period == 1 ~ "Post"
      ),
      levels = c("Pre", "Post")
    ),
    taxa    = factor(taxa),
    site    = relevel(factor(site), ref = "LTER1"),
    habitat = relevel(factor(habitat), ref = "Lagoon")
  )

# Quick verify
glimpse(recruit_transect)
unique(recruit_transect$habitat)

# Any out of bounds corals? 
# clean_filtered %>%
#   mutate(x = ifelse(x > 5, x / 10, x)) %>%
#   filter(x > 5) %>%
#   count(taxa, habitat, site)


# Check recruitment 
#view(recruit_transect)

write.csv(recruit_transect, file = "recruitment/data_recruitment/recruits_transect.csv")

```

\*\*\*\*Offshore for poc are recruits are counted in the first m\^2 (check and make sure this is the case, so not find recruits past the first meter sites)

```{r}
# Create filtered datasets for each taxa
dat_acr <- recruit_transect %>% filter(taxa == "Acr")
dat_poc <- recruit_transect %>% filter(taxa == "Poc")
dat_por <- recruit_transect %>% filter(taxa == "Por")

# Drop 2017 and only use complete cases

```



```{r}
# Acr
dat_acr %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  + 
    labs(title = "Habitat-wide Coral Recruitment of Acropora (2013-2024)", caption = "2017 not sampled.") + theme(legend.position = "none")


dat_por %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
    labs(title = "Habitat-wide Coral Recruitment of Porites (2013-2024)", caption = "2017 not sampled.") + theme(legend.position = "none")


dat_poc %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  +
    labs(title = "Habitat-wide Coral Recruitment of Pocillopera (2013-2024)", caption = "2017 not sampled.") 



```

## Model 1: Habitat level recruitment between bleaching periods

How does coral recruitment differ between lagoon and forereef habitats before and after the bleaching event?

-   **Formula: habitat \* year + (1\|site/transect_id)**
-   **site/transect_id: sites pooled as random... island-wide habitat question**
-   **year crossed random effect: absorbs annual variation**

\*\*SITE IS NOW A FIXED EFFECT TO CONTROL FOR GROUPING ALL TRANSECTS AND SITES TOGETHER (KEEP THEM DISTINCT!!)

\*\*Get estimate on link scale and then get se since only on link scale and then convert to response space

How did recruitment differ in the lagoon and forereef before and after bleaching? Recruitment... most abundant genera of corals (Por, Poc, Acr)... How you did it? (Poc and Por is for 1 m squared of forereef)

### Acropora

Changed to before and after comparison. ZiG: conditional is if there are no zeroes (ie. it was not there) and zero-inflated is with those zeroes (flexible). In order to interpret, will need to understand the predictions.

```{r}
# Show structural zeros or sparsity to show that tweedie did not work and ZIG is best

# For each transect what fraction of its years had ZERO recruits?
transect_zero_profile <- dat_acr_cc %>%
  group_by(transect_id, habitat) %>%
  summarize(
    n_years        = n(),
    n_zero_years   = sum(recruits_std == 0),
    prop_zero      = mean(recruits_std == 0),
    mean_recruits  = mean(recruits_std),
    max_recruits   = max(recruits_std),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_zero))

print(transect_zero_profile, n = 32)


# distribution of per-transect zero proportions
hist(transect_zero_profile$prop_zero, breaks = 10,
     main = "Per-transect proportion of zero years", xlab = "prop_zero")

# how many transects NEVER recruited at all?
transect_zero_profile %>%
  summarize(
    never_recruited   = sum(max_recruits == 0),     # 0 in all 11 years
    always_some       = sum(n_zero_years == 0),     # never had a zero year
    mixed             = sum(max_recruits > 0 & n_zero_years > 0)
  )
```

CHECK: 
Acropora recruitment zeros were strongly habitat-structured: 11 of the 13 transects with no recruits in ≥60% of survey years were located in the lagoon, including two lagoon transects (LTER5_BR_T01 and LTER5_BR_T02) that recorded no recruits throughout the entire 11-year study. In contrast, forereef transects recruited consistently, averaging 3–5 recruits per 5 m². The concentration of zeros within a habitat-defined subset of transects suggests a zero-generating process beyond sampling sparsity alone, supporting the use of a zero-inflated Gamma model with habitat included in the zero-inflation component.



```{r}
#| collapse: true
vars <- c("recruits_std","habitat","bleaching_status","site","transect_id","year")
dat_acr_cc <- dat_acr[complete.cases(dat_acr[, vars]), ]
nrow(dat_acr) - nrow(dat_acr_cc)   # check how many rows you lose

# Refit on complete cases (drop empty 2017 factor level)
dat_acr_cc <- dat_acr[complete.cases(dat_acr[, vars]), ]
dat_acr_cc$year <- droplevels(factor(dat_acr_cc$year))

q1_acr_bleaching_status <- glmmTMB(
  recruits_std ~ habitat*bleaching_status + site + (1 | transect_id) + (1|year),
  family    = ziGamma(link = "log"),
  #dispformula = ~habitat,
 # family = tweedie(link = "log"),
  ziformula = ~habitat*bleaching_status, # expect that the probability of a habitat experiencing complete recruitment failure (zero corals) changed after the bleaching event, and I expect it changed differently on the Forereef compared to the Lagoon
  data      = dat_acr_cc
)

summary(q1_acr_bleaching_status) 

```

```{r}
# emmeans
emmeans(q1_acr_bleaching_status, 
        ~ habitat * bleaching_status, 
        component = "response",     # <- Forces integration of the ZI part
        type = "response")
```


```{r}
# Year
acr_summary <- dat_acr_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(habitat, year, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

acr_q1_habitat_year_plot <- ggplot(
  acr_summary %>% filter(!is.na(bleaching_status)),
  aes(x = factor(year), y = mean_rec, fill = bleaching_status)
) +
  
  geom_col(width = 0.7, linewidth = 0.3) +
  
  geom_errorbar(
    aes(
      ymin = mean_rec,
      ymax = mean_rec + se_rec
    ),
    width = 0.25,
    linewidth = 0.6
  ) +
  
  facet_grid(
    ~ habitat,
    scales = "free",
    labeller = labeller(
      habitat = c(
        "BR" = "Lagoon",
        "OR" = "Forereef"
      )
    )
  ) +
  
  scale_fill_manual(
    values = c("dodgerblue", "red"),
    labels = c(
      "Pre-bleaching",
      "Post-bleaching"
    ),
    name = ""
  ) +
  
  scale_x_discrete(
    breaks = as.character(seq(2013, 2024))
  ) +
  
  theme_bw() +
  
  labs(
    x = "Year",
    y = expression("Mean recruits per 5 m"^2),
   # title = "Coral Recruitment of Acropora Between Habitats (2013–2024)",
    caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
  ) +
  theme(
  plot.background = element_rect(fill = "transparent", color = NA),
  legend.background = element_rect(fill = "transparent", color = NA),
  legend.box.background = element_rect(fill = "transparent", color = NA),
  legend.key = element_rect(fill = "transparent", color = NA),

  axis.title = element_text(size = 16),

  axis.text.x = element_text(
  size = 11,
  face = "plain",
  angle = 45,
  hjust = 1
),
  axis.text.y = element_text(size = 14, face = "plain"),

  strip.text = element_text(size = 18, face = "plain"),
  legend.text = element_text(size = 16, face = "plain")
)  
acr_q1_habitat_year_plot

ggsave(
  file = "/Users/vrs/MEDS/capstone/eds-coral-glmms/recruitment/figs/acr_q1_habitat_year_plot.png",
  acr_q1_habitat_year_plot,
  bg = "transparent",
  height = 8,
  width = 12
)


## blank
acr_q1_habitat_year_blank <- ggplot(
  acr_summary %>% filter(!is.na(bleaching_status)),
  aes(x = factor(year), y = mean_rec, fill = bleaching_status)
) +
  facet_grid(
    ~ habitat,
    scales = "free",
    labeller = labeller(
      habitat = c("BR" = "Lagoon", "OR" = "Forereef")
    )
  ) +
  scale_fill_manual(
    values = c("dodgerblue", "red"),
    labels = c("Pre-bleaching", "Post-bleaching"),
    name = ""
  ) +
  scale_x_discrete(breaks = c("2013", "2019", "2024")) +
  scale_y_continuous(limits = c(0, max(acr_summary$mean_rec + acr_summary$se_rec, na.rm = TRUE))) +
  theme_bw() +
  labs(
    x = "Year",
    y = expression("Mean recruits per 5 m"^2),
    caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
  ) +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.box.background = element_rect(fill = "transparent", color = NA),
    legend.key = element_rect(fill = "transparent", color = NA),
    axis.title = element_text(size = 16),
    axis.text.x = element_text(size = 11, face = "plain", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14, face = "plain"),
    strip.text = element_text(size = 18, face = "plain"),
    legend.text = element_text(size = 16, face = "plain")
  ) 

ggsave(
  file = "/Users/vrs/MEDS/capstone/eds-coral-glmms/recruitment/figs/acr_q1_habitat_year_blank.png",
  acr_q1_habitat_year_blank,
  bg = "transparent",
  height = 8,
  width = 12  # wider
)


```

```{r}
# total
# Year for all taxa
# Combine all three datasets
dat_all <- bind_rows(
  dat_acr_cc %>% mutate(taxa = "Acropora"),
  dat_poc_cc %>% mutate(taxa = "Pocillopora"),
  dat_por_cc %>% mutate(taxa = "Porites")
)

# Summarize by taxa and year
year_summary <- recruit_transect %>%
  filter(!is.na(recruits_std)) %>%
  group_by(taxa, year, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(
  year_summary %>% filter(!is.na(bleaching_status)),
  aes(x = factor(year), y = mean_rec, fill = taxa)
) +
  
  geom_col(width = 0.7, linewidth = 0.3) +
  
  geom_errorbar(
    aes(
      ymin = mean_rec,
      ymax = mean_rec + se_rec
    ),
    width = 0.25,
    linewidth = 0.6
  ) +
  
  facet_grid(
    ~ taxa,
    scales = "free") +
    # labeller = labeller(
    #   habitat = c(
    #     "BR" = "Lagoon",
    #     "OR" = "Forereef")
    
   
  
  # scale_fill_manual(
  #   values = c("dodgerblue", "red"),
  #   labels = c(
  #     "Pre-bleaching",
  #     "Post-bleaching"
  #   ),
  #   name = ""
  # ) +
  scale_x_discrete(
    breaks = as.character(seq(2013, 2024, by = 2))
  ) +
  
  theme_bw() +
  
  labs(
    x = "Year",
    y = expression("Mean recruits per 5 m"^2),
    title = "Coral Recruitment of Three Taxa (2013–2024)",
    caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
  ) +
  theme(
  plot.background = element_rect(fill = "transparent", color = NA),
  legend.background = element_rect(fill = "transparent", color = NA),
  legend.box.background = element_rect(fill = "transparent", color = NA),
  legend.key = element_rect(fill = "transparent", color = NA),

  axis.title = element_text(size = 16),

  axis.text.x = element_text(
  size = 11,
  face = "plain",
  angle = 45,
  hjust = 1
),
  axis.text.y = element_text(size = 14, face = "plain"),

  strip.text = element_text(size = 18, face = "plain"),
  legend.text = element_text(size = 16, face = "plain")
)  + geom_vline(
  xintercept = which(levels(factor(acr_summary$year)) == "2019"),
  linetype = "dashed",
  linewidth = 0.8,
  color = "black"
)



ggsave(
  file = "/Users/vrs/MEDS/capstone/eds-coral-glmms/recruitment/figs/acr_q1_habitat_year_plot.png",
  acr_q1_habitat_year_plot,
  bg = "transparent",
  height = 8,
  width = 7
)
```



```{r}
## habitat-bleaching
## Acropora
acr_summary <- dat_acr_cc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

acr_q1_habitat_bleaching_plot <- ggplot(acr_summary %>% filter(!is.na(bleaching_status)),
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Bleaching Event",
        labels = c(Pre = "Before", Post = "After")
    ) +
    theme_bw() +
    labs(
        x = "Period",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Acropora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) +
    theme(
        plot.background    = element_rect(fill = "transparent", color = NA),
        legend.background  = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill = "transparent", color = NA)
    )

acr_q1_habitat_bleaching_plot
ggsave(
    "recruitment/acr_q1_habitat_bleaching_plot.png",
    acr_q1_habitat_bleaching_plot,
    bg = "transparent"
)
```

### Pocillopera
site added for same reason max stated. ALSO, fails Levenne's test without it.

```{r}
# Poc
# q1_poc_bleaching_status <- glmmTMB(
#   recruits_std ~ habitat*bleaching_status + site + (1 | transect_id) + (1|year),
#   family = tweedie(link = "log"),
#  # family = ziGamma(link = "log"),
#   # ziformula = ~.,
#   data      = dat_poc,
#   na.action = na.exclude
# )
# 
# summary(q1_poc_bleaching_status)
vars <- c("recruits_std","habitat","bleaching_status","site","transect_id","year")
dat_poc_cc <- dat_poc[complete.cases(dat_poc[, vars]), ]
dat_poc_cc <- droplevels(dat_poc_cc)
nrow(dat_poc) - nrow(dat_poc_cc)   # likely 32 again, the missing-2017 rows

dat_poc_cc$year_num <- as.numeric(dat_poc_cc$year)
dat_poc_cc$year_fac <- as.factor(dat_poc_cc$year)

# q1_poc_bleaching_status <- glmmTMB(
#   recruits_std ~ habitat*bleaching_status + site + (1 | transect_id) + (1|year),
#   family = tweedie(link = "log"),
#   dispformula = ~ habitat, # Additive bc check levene's test (boxplots were red!)
#   data   = dat_poc_cc            # complete cases, no na.action
# )
# summary(q1_poc_bleaching_status)

q1_poc_bleaching_status <- glmmTMB(
  recruits_std ~ habitat * bleaching_status + site +
                 (1 | transect_id) + (1|year),
  family = tweedie(link = "log"),
 dispformula = ~ habitat,
  data   = dat_poc_cc           
)

summary(q1_poc_bleaching_status)

## Diagnose
res_poc1 <- simulateResiduals(q1_poc_bleaching_status, n = 1000)
plot(res_poc1)
testDispersion(res_poc1)
testZeroInflation(res_poc1)
diagnose(q1_poc_bleaching_status)
```


```{r}
# emmeans 
emmeans(q1_poc_bleaching_status, ~ habitat * bleaching_status, type = "response")
# add pairs for p val
emmeans(q1_poc_bleaching_status, ~ bleaching_status | habitat, type = "response") %>%
  pairs()

```


```{r}
## habitat-bleaching
poc_summary <- dat_poc_cc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(sum(!is.na(recruits_std))),        
        .groups = "drop"
    )

ggplot(poc_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    # scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Period",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Pocillopora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 
```

```{r}
# Add error bars
poc_summary <- dat_poc %>%
    #filter(!is.na(recruits_std)) %>%
    group_by(habitat, year, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(poc_summary,
       #%>% 
         #filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec ,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Pocillopora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) + theme(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA),
  legend.background = element_rect(fill = "transparent", colour = NA),
  legend.box.background = element_rect(fill = "transparent", colour = NA)
)
```

```{r}
# Habitat 
poc_summary <- dat_poc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(poc_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec - se_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m² (+SE)",
        title = "Habitat-wide Coral Recruitment of Pocillopora (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    )
```

Pocillopora recruitment differed significantly between the back reef and fore reef across the study period, with the fore reef recruiting significantly more overall (p \< 0.001). Back reef recruitment remained relatively stable between 2013 and 2019, with only 2018 exhibiting a notable increase (p = 0.035). A significant decline in recruitment followed by 2021 (p = 0.004).

The fore reef showed a striking decline in recruitment over the time period. The fore reef was the dominant recruitment habitat in 2013 but declined sharply relatively to the back reef from 2014 onwards (p = 0.028). This divergence in recruitment away from the back reef became increasingly significant through the post-bleaching year; every year from 2018-2024 showed a highly significant reduction in fore reef recruitment relative to 2013 (all p \< 0.001) except 2016 (p = 0.016). Recruitment was worse in the fore reef in 2022 (p \< 0.001) and did not show signs of significant recovery through 2024 (p \< 0.001).

#### 

```{r}
# Model validation 
## Model validation and assumptions 


# 1. for negative binomial, we will need to test for the overdispersion (qqplot), heteroscedacity, and zero-inflation

# 2. Residual diagnostics (outliers, non-normality, or deviations)
# Plot of predicted vs actual, should not see any curvature, will get noise but on the line mostly 

# Q2A (data never really matches the model... how different is the model from the assumption)
# switch to tweedie model for #Q2a.. try tweedie() 
```

### Porites

```{r}
# Por 
# q1_por_year <- glmmTMB(
#   recruits_std ~ habitat * factor(year) + site + (1 | transect_id),
#   family    = tweedie(link = "log"),
#   data      = dat_por,
#   na.action = na.exclude
# )
# 
# summary(q1_por_year)
dat_por_cc <- dat_por[complete.cases(dat_por[, vars]), ]
nrow(dat_por) - nrow(dat_por_cc)  

q1_por_bleaching_status <- glmmTMB(
  recruits_std ~ habitat*bleaching_status + site + (1 |transect_id) + (1|year),
  family = tweedie(link = "log"),
 # family = ziGamma(link = "log"),
  # ziformula = ~.,
 dispformula = ~ habitat,
  data      = dat_por_cc,
  #na.action = na.exclude
)
summary(q1_por_bleaching_status)

## Diagnose
res_por1<- simulateResiduals(q1_por_bleaching_status, n = 250)
plot(res_por1)
testDispersion(res_por1)
testZeroInflation(res_por1)
diagnose(q1_por_bleaching_status)
```

Both habitats declined by the same magnitude. 



```{r}
por_summary <- dat_por %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, year, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(por_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = factor(year), y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Year",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Porites Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    )
```

```{r}
# # Visualization
# # Observed data (keep all years including 2017)
# por_avg_recruits_year <- dat_por %>% 
#   group_by(habitat, year) %>% 
#   summarize(avg_recruit = mean(recruits_std, na.rm = TRUE)) %>% 
#   ungroup()
# 
# # Predictions (2017 excluded) 
# por_pred_grid_year <- dat_por %>%
#   filter(year != 2017) %>%
#   distinct(habitat, year)
# 
# por_pred_grid_year$predicted <- predict(q1_por_year, 
#                                         newdata = por_pred_grid_year,
#                                         type = "response",
#                                         re.form = NA)
# 
# # Join observed to predictions
# por_plot_data <- por_avg_recruits_year %>%
#   left_join(por_pred_grid_year, by = c("habitat", "year"))
# 
# # Plot
# ggplot(por_plot_data) + 
#   geom_col(aes(x = factor(year), y = avg_recruit, fill = habitat), 
#            alpha = 0.5) +
#   geom_point(aes(x = factor(year), y = predicted, col = habitat)) + 
#   geom_line(aes(x = factor(year), y = predicted, col = habitat, 
#                 group = habitat)) +  
#   facet_wrap(~ habitat) +
#   scale_fill_manual(
#     values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),  
#     name   = "Observed"
#   ) +
#   scale_color_manual(
#     values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"), 
#     name   = "Predicted"
#   ) +
#   scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
#   theme_bw() + 
#   theme( # Specify mean observed and predicted recruits
#     axis.text.x   = element_text(angle = 45, vjust = 1, hjust = 1),
#     legend.position = "bottom"
#   ) +
#   labs(
#     title   = "Habitat-wide Coral Recruitment of Porites (2013-2024)",
#     y       = "Recruits per 5 m²",
#     x       = "Year",
#     caption = "Bars = observed mean, points = model predicted. 2017 not sampled."
#   )
```

```{r}
## habitat-bleaching
por_summary <- dat_por_cc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(por_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    # scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Period",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Porites Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 
```

Porites recruitment was significantly higher in the back reef than the forereef across the entire study period (p \< 0.001), making it the only taxa to show back reef dominance. Back reef recruitment was exceptionally high in 2013 then declined sharply and significantly every subsequent year (all p \< 0.001), with the steepest drops occurring after 2019 and bottoming out in 2024, worst year on record (p \< 0.001). This monotonic decline suggests a long-term suppression of back reef Porites recruitment that predates the bleaching event and worsened dramatically after it.

The forereef told a different story. While it recruited significantly less than the back reef overall, the gap between habitats narrowed significantly in 2018 and 2019 (p = 0.021 and p = 0.037) as the back reef collapsed faster than the forereef during the bleaching period. By 2024 the interaction term reversed significantly (p = 0.009), not because the forereef recovered, but because the back reef had declined so severely that the two habitats converged near zero. This convergence reflects a system-wide collapse of Porites recruitment rather than any habitat-specific recovery, consistent with the near-zero values visible across both habitats in the post-2019 period.Model Validation

```{r}
## habitat-bleaching
por_summary <- dat_por_cc %>%
    filter(!is.na(recruits_std)) %>%
    group_by(habitat, bleaching_status) %>%  # add bleaching_status here
    summarise(
        mean_rec = mean(recruits_std, na.rm = TRUE),
        se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    )

ggplot(por_summary %>% filter(!is.na(bleaching_status)), 
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
    geom_col(width = 0.7, col = "white", linewidth = 0.3) +
    geom_errorbar(
        aes(ymin = mean_rec,
            ymax = mean_rec + se_rec),
        width = 0.25,
        linewidth = 0.6
    ) +
    facet_wrap(~ habitat) +
    scale_fill_manual(
        values = c(Pre = "dodgerblue", Post = "red"),
        name = "Period",
        labels = c(Pre = "Pre-bleaching (2013-2018)",
                   Post = "Post-bleaching (2019-2024)")
    ) +
    # scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
    theme_bw() +
    labs(
        x = "Period",
        y = "Mean recruits per 5 m²",
        title = "Coral Recruitment of Pocillopora Between Habitats (2013-2024)",
        caption = "Bars = observed mean, error bars = +SE. 2017 not sampled."
    ) 
```

### Model 1 Coefficient Tables

```{r}
models1 <- list(
  ACR = q1_acr_bleaching_status,
  POC = q1_poc_bleaching_status,
  POR = q1_por_bleaching_status
)

contrasts1_all <- imap_dfr(models1, ~ {
  emm <- emmeans(.x, ~ bleaching_status | habitat, type = "response")
  as.data.frame(contrast(emm, method = "pairwise", adjust = "bonferroni")) |>
    mutate(taxon = .y)
})

means1_all <- imap_dfr(models1, ~ {
  emm <- emmeans(.x, ~ bleaching_status | habitat, type = "response")
  as.data.frame(emm) |> mutate(taxon = .y)
})

means1_wide <- means1_all |>
  select(taxon, habitat, bleaching_status, response) |>
  tidyr::pivot_wider(names_from = bleaching_status, values_from = response) |>
  rename(Pre_mean = Pre, Post_mean = Post) |>
  mutate(across(c(Pre_mean, Post_mean), \(x) round(x, 2)))

contrasts1_all |>
  select(taxon, habitat, ratio, SE, z.ratio, p.value) |>
  left_join(means1_wide, by = c("taxon", "habitat")) |>
  mutate(
    across(c(ratio, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) |>
  select(taxon, habitat, Pre_mean, Post_mean, ratio, SE, z.ratio, p.value, sig) |>
  arrange(taxon, habitat) |>
  gt(groupname_col = "taxon") |>
  cols_label(
    habitat   = "Habitat",
    Pre_mean  = "Pre (recruits/5m²)",
    Post_mean = "Post (recruits/5m²)",
    ratio     = "Ratio (Pre/Post)",
    SE        = "SE",
    z.ratio   = "z",
    p.value   = "p",
    sig       = ""
  ) |>
  tab_header(title = "Pre- vs Post-bleaching recruitment contrasts by habitat")
```


```{r}
# Take 2
library(emmeans)
library(dplyr)
library(purrr)
library(gt)

# 1. Package your actual models into a named list
my_models <- list(
  ACR = q1_acr_bleaching_status, 
  POC = q1_poc_bleaching_status, 
  POR = q1_por_bleaching_status 
)

# 2. Extract the pairwise contrasts (Pre vs Post) within each habitat
rec_contrasts <- imap_dfr(my_models, ~ {
  as.data.frame(
    pairs(emmeans(.x, ~ bleaching_status | habitat, type = "response"))
  ) |> mutate(taxon = .y)
})

# 3. Format and render the stunning gt table
rec_contrasts |>
  select(taxon, habitat, contrast, ratio, SE, z.ratio, p.value) |>
  mutate(
    across(c(ratio, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 3, eps = 0.001)
  ) |>
  arrange(taxon, habitat) |>
  gt(groupname_col = "taxon") |>
  cols_label(
    habitat = "Habitat",
    contrast = "Contrast",
    ratio = "Response Ratio (Pre/Post)",
    z.ratio = "z",
    p.value = "p",
    sig = ""
  ) |>
  tab_header(
    title = "Pairwise Contrasts of Coral Recruitment Before and After 2019 Bleaching Event",
    subtitle = "Evaluated within Lagoon and Forereef habitats using GLMM contrasts"
  )
```





```{r}
contrasts |>
  select(taxon, habitat, contrast, ratio, SE, z.ratio, p.value) |>
  mutate(
    across(c(ratio, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) |>
  arrange(taxon, habitat) |>
  gt(groupname_col = "taxon") |>
  cols_label(
    habitat  = "Habitat",
    contrast = "Contrast",
    ratio    = "Ratio (Post/Pre)",
    z.ratio  = "z",
    p.value  = "p",
    sig      = ""
  ) |>
  tab_header(title = "Recruit density: Post/Pre-bleaching ratio by habitat")
```

```{r}
# Acropora
emtrends(q1_acr_bleaching_status, ~ habitat * bleaching_status, var = " ")
pairs(emtrends(q1_acr_bleaching_status, ~ habitat * period, var = "recruits_std"))

# Pocillopora
emtrends(m1_poc, ~ habitat * period, var = "log10_size_t0")
pairs(emtrends(m1_poc, ~ habitat * period, var = "log10_size_t0"))

# Porites
emtrends(m1_por, ~ habitat * period, var = "log10_size_t0")
pairs(emtrends(m1_por, ~ habitat * period, var = "log10_size_t0"))
```

```{r}
# Trial 2
models <- list(Acropora    = q1_acr_bleaching_status,
               Pocillopora = q1_poc_bleaching_status,
               Porites     = q1_por_bleaching_status)

results <- lapply(names(models), function(taxon) {
  emm <- emmeans(models[[taxon]], ~ bleaching_status | habitat,
                 type = "response")
  data.frame(taxon = taxon,
             as.data.frame(pairs(emm, reverse = TRUE)))
})
do.call(rbind, results)
```

### Model Validation

```{r}
# Try acropora
library(DHARMa)
acr_q1_res <- simulateResiduals(q1_acr_bleaching_status, n = 1000)
plot(acr_q1_res)                          # QQ + residual-vs-predicted
testDispersion(acr_q1_res)                # over/underdispersion
testZeroInflation(acr_q1_res)             # is the ZI part doing its job
testOutliers(acr_q1_res)

diagnose(q1_acr_bleaching_status)

## There is clear deviations, where? 
plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$habitat)

plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$bleaching_status)

plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$site)

plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$year)


####### Pocillopora
library(DHARMa)
poc_q1_res <- simulateResiduals(q1_poc_bleaching_status, n = 1000)
plot(poc_q1_res)                          # QQ + residual-vs-predicted
testDispersion(poc_q1_res)                # over/underdispersion
testZeroInflation(poc_q1_res)             # is the ZI part doing its job
testOutliers(poc_q1_res)

diagnose(q1_por_bleaching_status)

## There is clear deviations, where? 
plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$habitat)

plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$bleaching_status)

plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$site)

plotResiduals(acr_q1_res, form = model.frame(q1_acr_bleaching_status)$year)

```

### Model Takeaway

Emmeans will give percentage? look into this... or just the estimated emmeans + the pairs contrast for ratio 

Pre-bleaching, \[taxon\] recruitment was \[X\] per 5 m² in the lagoon and \[Y\] on the forereef. Post-bleaching it was \[X′\] and \[Y′\]. Recruitment changed by \[ratio→%\] in the lagoon (p = …) and \[ratio→%\] on the forereef (p = …); the habitat × bleaching interaction was \[significant/ns\] (p = …), indicating the lagoon–forereef difference \[did / did not\] change across the event.

## Model 2: Island-wide recruitment by site x habitat before and after bleaching

Do site × habitat differences in recruitment vary between pre and post-bleaching periods?

Formula: site \* habitat \* bleaching_status + (1\|transect_id) + (1\|year)

### Acropora

Transect_id is dropped here because every time transect_id was included it collapsed to essentially zero (1.367e-11, 1.431e-09)... means data couldn't distinguish between transects for Acropora. Sparse counts with many zeros meant there wasn't enough within-transect variation to estimate a transect-level random effect. Dropping it gave a stable and cleaner model.

```{r}
# AIC showed that dropping transect as random effect improved model ... variance was essentially 0
q2a_acr <- glmmTMB(
  recruits_std ~ site*habitat*bleaching_status +  (1 | year) ,
  # family = tweedie(link = "log"),
 family    = ziGamma(link = "log"),
 ziformula = ~habitat*bleaching_status, # go over why...
  data      = dat_acr_cc
#  na.action = na.exclude
)

#q2a_acr <- update(q2a_acr, ziformula = ~ habitat * bleaching_status)

summary(q2a_acr)
sim_acr <- simulateResiduals(q2a_acr, n = 1000)
plot(sim_acr)
plotResiduals(sim_acr, dat_acr_cc$habitat)

```


```{r}
# emmeans 
# Post/Pre within each site x habitat combination
emm_q2_acr <- emmeans(q2a_acr, ~ bleaching_status | site * habitat, component = "response", type = "response")

summary(emm_q2_acr)

## Before pairwise, it would be good to do predictions vs raw means for sanity check
acr_obs <- dat_acr_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(obs_mean = mean(recruits_std),
            n = n(), .groups = "drop")

emm_df <- as.data.frame(emm_q2_acr) %>%
  select(site, habitat, bleaching_status, emmean, asymp.LCL, asymp.UCL)

comparison <- emm_df %>%
  left_join(acr_obs, by = c("site", "habitat", "bleaching_status")) %>%
  mutate(diff = emmean - obs_mean)

print(comparison)
```



```{r}
# Prediction grid
acr_pred_grid_q2 <- dat_acr %>%
  distinct(site, habitat, bleaching_status)

acr_fit_q2 <- predict(q2a_acr, newdata = acr_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

acr_pred_grid_q2$predicted <- acr_fit_q2$fit
acr_pred_grid_q2$se <- acr_fit_q2$se.fit

# Observed means 
acr_obs_q2 <- dat_acr %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")

view(acr_obs_q2)
view(acr_pred_grid_q2)

```

```{r}
acr_summary_q2 <- dat_acr %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()), # se
    .groups = "drop"
  )


ggplot(acr_summary_q2,
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec,
                    ymax = mean_rec + se_rec),
                width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    labels = c(Pre = "Pre-bleaching", Post = "Post-bleaching")
  ) + 
  theme_bw() +
  labs(
    title = "Acropora Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Bleaching Period",
    y = "Mean recruits per 5 m²",
    fill = "Bleaching Period"
  ) + scale_x_discrete(labels = c(Pre = "Pre-bleaching\n(2013-2018)", 
                             Post = "Post-bleaching\n(2019-2024)")) +   theme(axis.title.x = element_blank())

```

### Pocillopora

\*NOTE: SIGNIFICANT OVERDISPERSION (0.02)... LIKELY FROM OUTLIERS... adding habitat+bleaching_status did not improve (check the within-group variance again)

```{r}
q2a_poc <- glmmTMB(
  recruits_std ~ site * habitat * bleaching_status + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
 dispformula = ~habitat,
  data      = dat_poc_cc,
 # na.action = na.exclude
)
summary(q2a_poc)

# predictions to get standard errors, conversion to response space!!! 

```


```{r}
# Post/Pre within each site x habitat combination
emm_q2_poc <- emmeans(q2a_poc, ~ bleaching_status | site * habitat, type = "response")

summary(emm_q2_poc)

## Before pairwise, it would be good to do predictions vs raw means for sanity check
poc_obs <- dat_poc_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(obs_mean = mean(recruits_std),
            n = n(), .groups = "drop")

emm_df_poc <- as.data.frame(emm_q2_poc) %>%
  select(site, habitat, bleaching_status, response, asymp.LCL, asymp.UCL)

comparison_poc_q2 <- emm_df_poc %>%
  left_join(poc_obs, by = c("site", "habitat", "bleaching_status")) %>%
  mutate(diff = response - obs_mean)

print(comparison_poc_q2)
```


```{r}
# Calculate the average observed recruits per transect at site-habitat-period level
poc_summary_q2 <- dat_poc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()), # se
    .groups = "drop"
  )


ggplot(poc_summary_q2,
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
  geom_col(width = 0.6, colour = "white") +
  geom_errorbar(aes(ymin = mean_rec,
                    ymax = mean_rec + se_rec),
                width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    labels = c(Pre = "Pre-bleaching", Post = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Pocillopora Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) + scale_x_discrete(labels = c(Pre = "Pre-bleaching\n(2013-2018)", 
                             Post = "Post-bleaching\n(2019-2024)")) +   theme(axis.text.x  = element_blank())

```

```{r}
# Prediction grid
poc_pred_grid_q2 <- dat_poc %>%
  distinct(site, habitat, bleaching_status)

poc_fit_q2 <- predict(q2a_poc, newdata = poc_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

poc_pred_grid_q2$predicted <- poc_fit_q2$fit
poc_pred_grid_q2$se <- poc_fit_q2$se.fit

# Observed means 
poc_obs_q2 <- dat_poc %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")
view(poc_obs_q2)
view(poc_pred_grid_q2)

# Combine 
# poc_q2_plot_df <- poc_pred_grid_q2 %>%
#   left_join(poc_obs_q2, by = c("site", "habitat", "bleaching_status")) %>%
#   pivot_longer(cols      = c(predicted, observed),
#                names_to  = "type",
#                values_to = "recruits") %>%
#   mutate(
#     bleaching_status = factor(bleaching_status, levels = c("Pre", "Post")),
#     type       = factor(type, levels = c("observed", "predicted")),
#     site       = factor(site, levels = c("LTER1", "LTER2", "LTER4", "LTER5"))
#   )


```

```{r}
# Predictions
ggplot() +
  # Observed bars
  geom_col(data = poc_obs_q2,
           aes(x = habitat, y = observed, fill = bleaching_status),
           position = position_dodge(width = .5), width = .5, col = "white", linewidth = 0.3, alpha = .7) +
  # # Predicted points
  # geom_point(data = poc_pred_grid_q2,
  #            aes(x = habitat, y = predicted, group = bleaching_status),
  #            position = position_dodge(width = .5),
  #            size = 3, shape = 18) +  
  # Predicted ci
  # geom_errorbar(
  #   data = poc_pred_grid_q2,
  #   aes(x    = habitat,
  #       ymin = exp(log(predicted) - 1.96 * (se / predicted)),
  #       ymax = exp(log(predicted) + 1.96 * (se / predicted)),
  #       group = bleaching_status),
  #   position = position_dodge(width = 0.75),
  #   width = 0.25, linewidth = 0.6
  # ) +
  facet_wrap(~ site) +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    name   = "Period",
    labels = c(Pre = "Pre-bleaching (2013-2018)",
               Post = "Post-bleaching (2019-2024)")
  ) +
  labs(
    title    = "Pocillopora Island-Wide Recruitment",
    caption = "Bars = observed mean, point = model predicted",
    x        = "Habitat",
    y        = "Recruits per 5 m²"
  ) +
  theme_bw() +
  theme(
    legend.position    = "bottom"
  )



```

Prior to bleaching, Pocillopora recruitment was significantly higher on the fore reef than the back reef island-wide (p = .008). There were no significant differences among sites within either habitat, suggesting that there is a consistent habitat-level pattern across Mo'orea. Back reef recruitment remained low and stable across all four sites pre-bleaching while fore reef recruitment was higher at all sites relative to their respective back reefs.

Recruitment patterns in Pocillopora shifted significantly in both habitats. Fore reef recruitment declined sharply at both sites, and the previous advantage for recruitment held in the fore reef disappeared at LTER1 (p \< 0.001). Recruitment also declined significantly at the LTER 2 back reef (p = 0.049). However, the degree of this decline varied among sites. LTER 2 fore reef experienced a smaller decline than LTER1 (p = 0.031). LTER 5 fore reef showed no significant decline post-bleaching, maintaining pre-bleaching recruitment levels (p \< 0.001). Overall, bleaching reduced fore reef recruitment across the island, but some sites (especially LTER 5) appeared to serve as a refuge that could potentially help Pocillopora recovery.

### Porites

Outliers observed in 2013

```{r}
q2a_por <- glmmTMB(
  recruits_std ~ site * habitat * bleaching_status + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  dispformula = ~habitat,
  data      = dat_por_cc,
  #na.action = na.exclude
)
summary(q2a_por)

```

```{r}
# Prediction grid
por_pred_grid_q2 <- dat_por %>%
  distinct(site, habitat, bleaching_status)

por_fit_q2 <- predict(q2a_por, newdata = por_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

por_pred_grid_q2$predicted <- por_fit_q2$fit
por_pred_grid_q2$se <- por_fit_q2$se.fit

# Observed means 
por_obs_q2 <- dat_por %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")
view(por_obs_q2)
view(por_pred_grid_q2)
```

```{r}
# Post/Pre within each site x habitat combination
emm_q2_por <- emmeans(q2a_por, ~ bleaching_status | site * habitat, type = "response")

summary(emm_q2_por)

## Before pairwise, it would be good to do predictions vs raw means for sanity check
por_obs <- dat_por_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(obs_mean = mean(recruits_std),
            n = n(), .groups = "drop")

emm_df_por <- as.data.frame(emm_q2_por) %>%
  select(site, habitat, bleaching_status, response, asymp.LCL, asymp.UCL)

comparison_por_q2 <- emm_df_por %>%
  left_join(por_obs, by = c("site", "habitat", "bleaching_status")) %>%
  mutate(diff = response - obs_mean)

print(comparison_por_q2)
```



```{r}
outliers(res_q2a_por)
dat_por_cc[outliers(res_q2a_por), ]
```

```{r}
# Calculate the average observed recruits per transect at site-habitat-period level
por_summary_q2 <- dat_por_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, bleaching_status) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec = sd(recruits_std, na.rm = TRUE) / sqrt(n()), # se
    .groups = "drop"
  )


ggplot(por_summary_q2,
       aes(x = bleaching_status, y = mean_rec, fill = bleaching_status)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec,
                    ymax = mean_rec + se_rec),
                width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c(Pre = "dodgerblue", Post = "red"),
    labels = c(Pre = "Pre-bleaching", Post = "Post-bleaching")
  ) +
  theme_bw() + facet_grid(habitat~site) + 
  labs(
    title = "Porites Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Bleaching Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) + scale_x_discrete(labels = c(Pre = "Pre-bleaching\n(2013-2018)", 
                             Post = "Post-bleaching\n(2019-2024)"))
```

### Model Diagnostics

```{r}
res_q2a_acr <- simulateResiduals(q2a_acr, n = 250)
plot(res_q2a_acr)
testDispersion(res_q2a_acr)
testZeroInflation(res_q2a_acr)
testOutliers(res_q2a_acr)

res_q2a_poc <- simulateResiduals(q2a_poc, n = 1000)
plot(res_q2a_poc)
testDispersion(res_q2a_poc)
testZeroInflation(res_q2a_poc)

res_q2a_por <- simulateResiduals(q2a_por, n = 1000)
plot(res_q2a_por)
testDispersion(res_q2a_por)
testZeroInflation(res_q2a_por)
```

### Model 2 Emmeans
```{r}
## Now do pairwise
acr_contrasts <- pairs(emmeans(q2a_acr, ~ bleaching_status | site * habitat,
                               component = "response", type = "response"),
                       reverse = TRUE) %>%
  as.data.frame() %>%
  mutate(taxa = "Acropora")

# same for poc and por once those emmeans are built
poc_contrasts <- pairs(emmeans(q2a_poc, ~ bleaching_status | site * habitat,
                               component = "response", type = "response"),
                       reverse = TRUE) %>%
  as.data.frame() %>%
  mutate(taxa = "Pocillopora")

por_contrasts <- pairs(emmeans(q2a_por, ~ bleaching_status | site * habitat,
                               component = "response", type = "response"),
                       reverse = TRUE) %>%
  as.data.frame() %>%
  mutate(taxa = "Por")

# Build
all_contrasts <- bind_rows(acr_contrasts, poc_contrasts, por_contrasts)
all_contrasts
```

```{r}
# Build gt table
library(gt)

q2_table <- all_contrasts %>%
  mutate(
    taxa = recode(taxa, "Por" = "Porites"),
    sig  = ifelse(p.value < 0.05, "Significant", "n.s."),
    direction = case_when(
      p.value >= 0.05 ~ "No change",
      estimate < 0    ~ "Decline",
      estimate > 0    ~ "Increase"
    )
  ) %>%
  select(taxa, site, habitat, estimate, SE, p.value, direction) %>%
  arrange(taxa, habitat, site) %>%
  gt(groupname_col = "taxa") %>%
  tab_header(
    title = "Post-bleaching change in coral recruitment by site and habitat",
    subtitle = "Estimated change (Post-Pre) in recruits per 5 m², from site × habitat × bleaching GLMMs"
  ) %>%
  fmt_number(columns = c(estimate, SE), decimals = 2) %>%
  fmt_number(columns = p.value, decimals = 3) %>%
  cols_label(
    site = "Site",
    habitat = "Habitat",
    estimate = "Change (Post − Pre)",
    SE = "SE",
    p.value = "p-value",
    direction = "Direction"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = direction, rows = p.value < 0.05)
  ) %>%
  # data_color(
  #   columns = direction,
  #   fn = function(x) dplyr::case_when(
  #     x == "Decline"   ~ "#FAECE7",
  #     x == "Increase"  ~ "#E6F1FB",
  #     x == "No change" ~ "#F1EFE8",
  #     TRUE ~ "#FFFFFF"
  #   )
 # ) %>%
  tab_footnote(
    footnote = "Values are absolute changes on the response scale (recruits per 5 m²). Magnitudes reflect each taxa's pre-bleaching baseline and are not directly comparable across taxa.",
    locations = cells_column_labels(columns = estimate)
  ) %>%
  tab_source_note(
    source_note = "Estimates are response-scale changes (Post-Pre, recruits per 5 m²) for all taxa. Models: zero-inflated Gamma for Acropora, Tweedie for Pocillopora and Porites. Significance at p < 0.05."
  )

q2_table

# Export 
library(gt)
gtsave(q2_table, "recruitment/figs/q2_table.docx")

```

