---
title: "Recruitment GLMM"
format: html
editor_options:
  chunk_output_type: console
---

# Overview

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following questions were utilized to guide the analysis:

1.  **How did demographic rates vary across habitats before and after the 2019 bleaching event?**
2.  **How did demographic rates vary across both site and habitat before and after the 2019 bleaching event?**

**Reference group for all models:** Lagoon (Model 1) or LTER1 lagoon (Model 2), before bleaching.

# Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(glmmTMB)
library(ggeffects)
library(DHARMa)
library(emmeans)
library(dplyr)
library(purrr)
library(gt)
```

# Data Preparation

Load the cleaned data (referenced from a separate repository).

```{r}
#| message: false
#| warning: false
coral <- read_csv(here::here("..", "eds-coral-data-storage-management", "analysisdata", "coral_demography_analysis_data.csv"), guess_max = 10000) # Corrects column types

glimpse(coral)
```

In case of habitats being coded as `BR` and `OR` in the data, the following can be recoded as lagoon and forereef, respectively.

```{r}
coral$habitat <- fct_recode(coral$habitat, "Lagoon" = "BR","Forereef" = "OR")
```

Some transects were duplicate records of existing transects under slightly different naming conventions. The `transect_id` values were standardized so duplicate transects merged with their primary IDs.

```{r}
clean_coral <- coral %>%
  mutate(transect_id = case_when(
    transect_id == "LTER1_OR_TO1" ~ "LTER1_OR_T01",
    transect_id == "LTER1_OR_T02_2021" ~ "LTER1_OR_T02",
    transect_id == "LTER2_OR_T04_OLD" ~ "LTER2_OR_T04",
    transect_id == "LTER5_OR_T02_OLD" ~ "LTER5_OR_T02",
    transect_id == "LTER5_OR_T03_PRE2020" ~ "LTER5_OR_T03",
    TRUE ~ transect_id
  ))

unique(clean_coral$transect_id)
```

*Pocillopora* recruits on the forereef were sampled within the first 1 m² of each 5 m² transect as per the sampling protocol. A total of 26 *Pocillopora* recruit observations (2.9% of forereef *Pocillopora* recruits) recorded beyond the 1 m² boundary were excluded as out-of-bounds. These exclusions were confirmed with the field data collector prior to analysis.

```{r}
# Check how many forereef Poc recruits remain after the 1 m^2 filter
poc_before <- clean_coral %>%
  filter(taxa == "Poc", habitat == "Forereef", dyn_recruitment == 1) %>%
  nrow()

poc_after <- clean_coral %>%
  filter(taxa == "Poc", habitat == "Forereef", dyn_recruitment == 1, x <= 1) %>%
  nrow()

cat("Poc Forereef recruits before filter:", poc_before, "\n")
cat("Poc Forereef recruits after filter:", poc_after, "\n")
cat("Recruits removed:", poc_before - poc_after, "\n")
cat("Retention rate:", round(100 * poc_after / poc_before, 1), "%\n")
```

## Recruitment Standardization

Standardization was necessary to ensure an apples-to-apples comparison across surveys. Because forereef *Pocillopora* recruits were counted over 1 m² while all other observations were counted over 5 m², forereef *Pocillopora* counts were multiplied by 5 to express all recruitment on a common per 5 m² scale. Recruits were summed per transect, year, and period. 2017 was set to `NA` as it was recorded as non-sampling year. *Millepora* and out-of-bounds forereef *Pocillopora* recruits were excluded prior to standardization.

```{r}
recruit_transect <- clean_coral %>%
  filter(
    !(habitat == "Forereef" & taxa == "Poc" & x > 1 & dyn_recruitment == 1),# out-of-bounds Poc
    taxa != "Mil" # drop Millepora
  ) %>%
  group_by(taxa, site, habitat, transect_id, year, period) %>%
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    n_recruits = ifelse(year == 2017, NA, n_recruits), # 2017 set to NA
    recruits_std = case_when(
      habitat == "Forereef" & taxa == "Poc" ~ n_recruits * 5, # scale 1 m^2 to 5 m^2
      TRUE ~ n_recruits  # already per 5 m^2
    ),
    period = factor(period), # convert to factor for model purposes
    taxa    = factor(taxa),
    site    = relevel(factor(site), ref = "LTER1"),     # LTER1 = reference site
    habitat = relevel(factor(habitat), ref = "Lagoon")  # Lagoon = reference habitat
  )

glimpse(recruit_transect)
```

Filter to per-taxa datasets and only include complete cases (ie. drop the empty 2017 factor level).

```{r}
dat_acr <- recruit_transect %>% filter(taxa == "Acr")
dat_poc <- recruit_transect %>% filter(taxa == "Poc")
dat_por <- recruit_transect %>% filter(taxa == "Por")

vars <- c("recruits_std", "habitat", "period", "site", "transect_id", "year")

dat_acr_cc <- dat_acr[complete.cases(dat_acr[, vars]), ]
dat_acr_cc$year <- droplevels(factor(dat_acr_cc$year))

dat_poc_cc <- dat_poc[complete.cases(dat_poc[, vars]), ]
dat_poc_cc$year <- droplevels(factor(dat_poc_cc$year))

dat_por_cc <- dat_por[complete.cases(dat_por[, vars]), ]
dat_por_cc$year <- droplevels(factor(dat_por_cc$year))
```

# Model 1: Habitat-Level Recruitment Before and After Bleaching

**Question:** Does recruitment differ between forereef and lagoon habitats before and after the 2019 bleaching event?

-   **Fixed effects:** `habitat * period + site`. The `habitat * period` interaction tests whether the bleaching effect differs between habitats. Site is included as an additive fixed effect to keep the four sites distinct and control for site-level differences in baseline recruitment without modeling site-specific bleaching responses.

-   **Random effects:** `(1 | transect_id)` for repeated sampling of transects, `(1 | year)` for inter-annual variation.

-   **Distribution:** Zero-inflated Gamma for *Acropora*; Tweedie for *Pocillopora* and *Porites* (both log link).

-   **Reference group:** Lagoon, before bleaching.

## Fit models

### *Acropora*

Modeled with a zero-inflated Gamma. The conditional (Gamma) component models recruitment abundance where recruitment occurred; the zero-inflation component (`ziformula = ~ habitat * period`) models the probability of a transect having no recruits, allowing it to vary by habitat and period. Site was included as a fixed effect because it improved fit (AIC difference = 7) and diagnostics.

```{r}
coral_acr_hab <- glmmTMB(
  recruits_std ~ habitat * period + site + (1 | transect_id) + (1 | year),
  family    = ziGamma(link = "log"),
  ziformula = ~ habitat * period,
  data = dat_acr_cc,
  na.action = na.exclude
)

summary(coral_acr_hab)
```

### *Pocillopora*

Modeled with a Tweedie distribution. Dispersion was modeled as a function of habitat (`dispformula = ~ habitat`) because forereef recruitment was much more variable than lagoon recruitment. This approach improved residual and dispersion diagnostics. Site was included as a fixed effect.

```{r}
coral_poc_hab <- glmmTMB(
  recruits_std ~ habitat * period + site + (1 | transect_id) + (1 | year),
  family      = tweedie(link = "log"),
  dispformula = ~ habitat,
  data  = dat_poc_cc
)

summary(coral_poc_hab)
```

### *Porites*

Modeled with a Tweedie distribution, with dispersion modeled as a function of habitat.

```{r}
coral_por_hab <- glmmTMB(
  recruits_std ~ habitat*period +  site + (1 |transect_id) + (1|year),
  family = tweedie(link = "log"),
  dispformula = ~ habitat,
  na.action = na.exclude,
  data = dat_por_cc,
)

summary(coral_por_hab)
```

## Visualize models (observed means)

Bars show observed mean recruitment per 5 m², and error bars are +1 SE. 2017 was not sampled.

### *Acropora*

```{r}
# Acropora observed means
acr_summary_hab <- dat_acr_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(habitat, period) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec   = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

acr_plot_hab <- ggplot(acr_summary_hab,
       aes(x = factor(period), y = mean_rec, fill = factor(period))) +  
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec, ymax = mean_rec + se_rec), width = 0.25) +
  facet_grid(~habitat, scales = "free_y") +
  scale_fill_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Acropora Habitat-Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    caption = "Bars = observed mean, error bars = +SE. 2017 not sampled.",
    fill = "Period"
  ) +
  scale_x_discrete(labels = c("0" = "Pre", "1" = "Post")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8),
        title = element_text(size = 9.5), 
        plot.caption = element_text(size = 7))

acr_plot_hab

ggsave(
    "recruitment/figs/acr_recruitment_hab_plot.png",
    acr_plot_hab, 
    width = 6.5, 
    height = 4,
    dpi = 300
)
```

### *Pocillopora*

```{r}
# Pocillopora observed means 
poc_summary_hab <- dat_poc_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(habitat, period) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec   = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

poc_plot_hab <- ggplot(poc_summary_hab,
       aes(x = factor(period), y = mean_rec, fill = factor(period))) +  
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec, ymax = mean_rec + se_rec), width = 0.25) +
  facet_grid(~habitat, scales = "free_y") +
  scale_fill_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Pocillopora Habitat-Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    caption = "Bars = observed mean, error bars = +SE. 2017 not sampled.",
    fill = "Period"
  ) +
  scale_x_discrete(labels = c("0" = "Pre", "1" = "Post")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8),
        title = element_text(size = 9), 
        plot.caption = element_text(size = 7))

poc_plot_hab

ggsave(
    "recruitment/figs/poc_recruitment_hab_plot.png",
    poc_plot_hab, 
    width = 6.5, 
    height = 4,
    dpi = 300
)
```

### *Porites*

```{r}
# Porites observed means 
por_summary_hab <- dat_por_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(habitat, period) %>%
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec   = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

por_plot_hab <- ggplot(por_summary_hab,
       aes(x = factor(period), y = mean_rec, fill = factor(period))) +  
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec, ymax = mean_rec + se_rec), width = 0.25) +
  facet_grid(~habitat, scales = "free_y") +
  scale_fill_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Porites Habitat-Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    caption = "Bars = observed mean, error bars = +SE. 2017 not sampled.",
    fill = "Period"
  ) +
  scale_x_discrete(labels = c("0" = "Pre", "1" = "Post")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8),
        title = element_text(size = 9), 
        plot.caption = element_text(size = 7))

por_plot_hab

ggsave(
    "recruitment/figs/por_recruitment_hab_plot.png",
    por_plot_hab, 
    width = 6.5, 
    height = 4,
    dpi = 300
)
```

## Model Checks

### *Acropora* Residuals

```{r}
# Acropora
sim_acr_hab <- simulateResiduals(coral_acr_hab, n = 1000)
plot(sim_acr_hab)
```

### *Pocillopora* Residuals

```{r}
# Pocillopora
sim_poc_hab <- simulateResiduals(coral_poc_hab, n = 1000)
plot(sim_poc_hab)
```

### *Porites* Residuals

```{r}
# Porites
sim_por_hab <- simulateResiduals(coral_por_hab, n = 1000)
plot(sim_por_hab)
```

# Model 2: Island-Wide Recruitment by Site × Habitat Before/After Bleaching

**Question:** Does recruitment differ across all site-habitat combinations before and after the 2019 bleaching event (island-wide)?

-   **Fixed effects:** `site * habitat * period` (full three-way interaction).

-   **Random effects:** `(1 | year)` for between-year variation. Transect was dropped for *Acropora* because its variance collapsed to zero; however, this was retained for the other taxa.

-   **Distribution:** Zero-inflated Gamma for *Acropora*. Tweedie for *Pocillopora* and *Porites*.

For the *Acropora* zero-inflation component, the three-way interaction caused separation and inflated SEs, likely due to some site\*habitat\*period cells being either all-zero or non-zero. The zero-inflation formula was simplified to `~ habitat * period + site` where the bleaching's effect on the probability of a transect having zero recruits was allowed to differ by habitat, while site shifts the baseline zero probability.

**Reference group:** LTER1 lagoon, before bleaching.

## Fit models

### *Acropora*

```{r}
coral_acr_hab_site <- glmmTMB(
  recruits_std ~ site * habitat * period + (1 | year),
  family = ziGamma(link = "log"),
  ziformula = ~ habitat * period + site,
  data  = dat_acr_cc
)

summary(coral_acr_hab_site)
```

### *Pocillopora*

```{r}
coral_poc_hab_site <- glmmTMB(
  recruits_std ~ site * habitat * period + (1 | transect_id) + (1 | year),
  family = tweedie(link = "log"),
 dispformula = ~habitat, # Handles within-variance best (interacting with site or period made simulated residuals significant and worsened the dispersion), term is significant
  data = dat_poc_cc,
)

summary(coral_poc_hab_site)
```

### *Porites*

```{r}
coral_por_hab_site <- glmmTMB(
  recruits_std ~ site * habitat * period + (1 | transect_id) + (1 | year),
  family = tweedie(link = "log"),
 # dispformula = ~habitat, # not needed since no significant within-group variance
  data = dat_por_cc
 )

summary(coral_por_hab_site)
```

## Visualize Models (observed means)

Observed mean recruitment per 5 m² by site and habitat. Error bars are +1 SE. Note differing y-axis scales between habitat rows.

### *Acropora*

```{r}
# Acropora site-habitat observed means 
acr_summary_hab_site <- dat_acr_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, period) %>% 
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec   = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

acr_plot_hab_site <- ggplot(acr_summary_hab_site,
       aes(x = factor(period), y = mean_rec, fill = factor(period))) + 
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec, ymax = mean_rec + se_rec), width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Acropora Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) +
  scale_x_discrete(labels = c("0" = "Pre", "1" = "Post")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8),
        title = element_text(size = 8), 
        plot.caption = element_text(size = 7))

acr_plot_hab_site

ggsave(
    "recruitment/figs/acr_recruitment_hab_site_plot.png",
    acr_plot_hab_site, 
    width = 6.5, 
    height = 4,
    dpi = 300
)
```

### *Pocillopora*

```{r}
# Pocillopora site-habitat observed means 
poc_summary_hab_site <- dat_poc_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, period) %>% 
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec   = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

poc_plot_hab_site <- ggplot(poc_summary_hab_site,
       aes(x = factor(period), y = mean_rec, fill = factor(period))) + 
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec, ymax = mean_rec + se_rec), width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Pocillopora Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) +
  scale_x_discrete(labels = c("0" = "Pre", "1" = "Post")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8),
        title = element_text(size = 8), 
        plot.caption = element_text(size = 7))

poc_plot_hab_site

ggsave(
    "recruitment/figs/poc_recruitment_hab_site_plot.png",
    poc_plot_hab_site, 
    width = 6.5, 
    height = 4,
    dpi = 300
)
```

### *Porites*

```{r}
# Porites site-habitat observed means 
por_summary_hab_site <- dat_por_cc %>%
  filter(!is.na(recruits_std)) %>%
  group_by(site, habitat, period) %>% 
  summarise(
    mean_rec = mean(recruits_std, na.rm = TRUE),
    se_rec   = sd(recruits_std, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

por_plot_hab_site <- ggplot(por_summary_hab_site,
       aes(x = factor(period), y = mean_rec, fill = factor(period))) + 
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = mean_rec, ymax = mean_rec + se_rec), width = 0.25) +
  facet_grid(habitat ~ site, scales = "free_y") +
  scale_fill_manual(
    values = c("0" = "dodgerblue", "1" = "red"),
    labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")
  ) +
  theme_bw() +
  labs(
    title = "Porites Site-Habitat Wide Differences in Recruitment Before and After Bleaching",
    x = "Period",
    y = "Mean recruits per 5 m²",
    fill = "Period"
  ) +
  scale_x_discrete(labels = c("0" = "Pre", "1" = "Post")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8),
        title = element_text(size = 8), 
        plot.caption = element_text(size = 7))

por_plot_hab_site

ggsave(
    "recruitment/figs/por_recruitment_hab_site_plot.png",
    por_plot_hab_site, 
    width = 6.5, 
    height = 4,
    dpi = 300
)
```

## Model Checks

### *Acropora* Residuals

```{r}
# Acropora
sim_acr_hab_site <- simulateResiduals(coral_acr_hab_site, n = 1000)
plot(sim_acr_hab_site)
```

### *Pocillopora* Residuals

```{r}
# Pocillopora
sim_poc_hab_site <- simulateResiduals(coral_poc_hab_site, n = 1000)
plot(sim_poc_hab_site)
```

### *Porites* Residuals

```{r}
# Porites
sim_por_hab_site <- simulateResiduals(coral_por_hab_site, n = 1000)
plot(sim_por_hab_site)
```

## Model Contrasts

### Model 1 Contrast Table

Estimated mean differences in recruitment by habitat, averaged over sites.

```{r}
# Create list of models
models1 <- list(
  Acr = coral_acr_hab,
  Poc = coral_poc_hab,
  Por = coral_por_hab
)

# Create function to report contrasts for each model
get_contrasts_q1 <- function(model, taxa) {
  emmeans(model, ~ period | habitat, component = "response") %>% # component= response for Acr ZIG (two models)
    pairs(reverse = TRUE) %>%
    summary(infer = TRUE) %>%
    as.data.frame() %>%
    rename_with(~ "lower.CL", any_of(c("asymp.LCL", "lower.HPD", "lower.CL"))) %>%
    rename_with(~ "upper.CL", any_of(c("asymp.UCL", "upper.HPD", "upper.CL"))) %>%
    mutate(
      taxa = taxa, # match function input for taxa
      contrast = "Post - Pre" # overwrite period contrasts
    ) %>%
    dplyr::select(taxa, habitat, contrast, estimate, lower.CL, upper.CL, SE, z.ratio, p.value)
}

contrasts_q1 <- imap_dfr(models1, ~ get_contrasts_q1(.x, .y))

contrasts_q1 %>%
  mutate(
    across(c(estimate, lower.CL, upper.CL, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      # define p-value ranges
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) %>%
  arrange(taxa, habitat) %>%
  gt(groupname_col = "taxa") %>%
  cols_label(
    habitat  = "Habitat",
    contrast = "Contrast",
    estimate = "Recruit Difference",
    z.ratio  = "z",
    p.value  = "p",
    sig      = ""
  ) %>%
  tab_header(title = "Pairwise Differences of Coral Recruitment Following the 2019 Bleaching Event")
```

### Model 2 Contrasts Table

Estimated mean differences (post-pre) in recruitment by site and habitat. Negative values indicate a post-bleaching decline.

```{r}
# Create list of models
models2 <- list(
  Acr = coral_acr_hab_site,
  Poc = coral_poc_hab_site,
  Por = coral_por_hab_site
)

# Create function to report contrasts for each model
get_contrasts_q2 <- function(model, taxa) {
  emmeans(model, ~ period | site * habitat, component = "response") %>% # component = response for Acr ZIG (two models)
    pairs(reverse = TRUE) %>%
    summary(infer = TRUE) %>%
    as.data.frame() %>%
    rename_with(~ "lower.CL", any_of(c("asymp.LCL", "lower.HPD", "lower.CL"))) %>%
    rename_with(~ "upper.CL", any_of(c("asymp.UCL", "upper.HPD", "upper.CL"))) %>%
    mutate(
      taxa = taxa, # match function input for taxa
      contrast = "Post - Pre"  # overwrite period contrasts
    ) %>%
    dplyr::select(taxa, site, habitat, contrast, estimate, lower.CL, upper.CL, SE, z.ratio, p.value)
}

all_contrasts_2 <- imap_dfr(models2, ~ get_contrasts_q2(.x, .y))

all_contrasts_2 %>%
  mutate(
    across(c(estimate, lower.CL, upper.CL, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      # define p-value ranges
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) %>%
  arrange(taxa, site, habitat) %>%
  gt(groupname_col = "taxa") %>%
  cols_label(
    site     = "Site",
    habitat  = "Habitat",
    contrast = "Contrast",
    estimate = "Recruit Difference",
    z.ratio  = "z",
    p.value  = "p",
    sig      = ""
  ) %>%
  tab_header(title = "Pairwise Differences of Coral Recruitment by Site and Habitat Following the 2019 Bleaching Event")
```
