---
title: "recruitment_glmm_walkthrough"
format: html
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora from 2013-2024?**

## Data Preparation

### Load and Clean

The following packages were required prior to conducting additional data wrangling:

```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation 
```

Then, the cleaned data (referenced from a separate repository) was loaded in.

```{r}
#| message: false
#| warning: true
# clean_coral <- read_csv( 
# "~/MEDS/capstone/eds-coral-data-storage-management/analysisdata/coral_demography_analysis_data.csv", guess_max = 1000)

clean_coral <- read_csv( 
"~/MEDS/capstone/eds-coral-glmms/recruitment/coral_demography_analysis_data_vedika_copy.csv", guess_max = 1000 )

# Take a look at the cleaned coral dataset
glimpse(clean_coral)

```
```{r}
# Rename habitats 
clean_coral$habitat <- fct_recode(clean_coral$habitat, 
                "Back Reef" = "BR", 
                "Fore Reef" = "OR")
```

Standardize recruitment for every meters squared on the POC on the forereef... CHECK!! 

```{r}
# recruit_transect <- clean_coral %>%
#   group_by(taxa, site, habitat, transect_id, year) %>%
#   summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
#   mutate(
#     n_recruits = ifelse(year == 2017, NA, n_recruits),
#     recruits_std = n_recruits / 5,
#     recruits_int = as.integer(round(recruits_std)),
#     pre_post_f = factor(
#       case_when(
#         year >= 2013 & year <= 2018 ~ "Pre",
#         year >= 2019 & year <= 2024 ~ "Post"
#       ),
#       levels = c("Pre", "Post")
#     ),
#     taxa    = factor(taxa),
#     site    = relevel(factor(site),    ref = "LTER1"),
#     habitat = relevel(factor(habitat), ref = "BR")
#   )

recruit_transect <- clean_coral %>%
  filter(x <= 5) %>%  # Keep transect corals only
  group_by(taxa, site, habitat, transect, year) %>%
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(n_recruits = ifelse(year == 2017, NA, n_recruits)) %>%  # NA before standardizing
  mutate(
    recruits_std = n_recruits / 5,
    recruits_int = as.integer(round(recruits_std)),
    bleaching_status   = factor(
      case_when(
        year >= 2013 & year <= 2018 ~ "Pre",
        year >= 2019 & year <= 2024 ~ "Post"
      ),
      levels = c("Pre", "Post")
    ),
    taxa    = factor(taxa),
    site    = relevel(factor(site),    ref = "LTER1"),
    habitat = relevel(factor(habitat), ref = "Back Reef"),
    transect_id = paste(site, habitat, transect, sep = "_"),
    transect_id = case_when(
      transect_id == "LTER1_OR_TO1"         ~ "LTER1_OR_T01",
      transect_id == "LTER1_OR_T02_2021"    ~ "LTER1_OR_T02",
      transect_id == "LTER2_OR_T04_OLD"     ~ "LTER2_OR_T04",
      transect_id == "LTER5_OR_T02_OLD"     ~ "LTER5_OR_T02",
      transect_id == "LTER5_OR_T03_PRE2020" ~ "LTER5_OR_T03",
      TRUE ~ transect_id
    )
  )

# Verify
unique(recruit_transect$transect_id)
#unique(clean_coral$transect_id)
view(recruit_transect)
```

****Offshore for poc are recruits are counted in the first m^2 (check and make sure this is the case, so not find recruits past the first meter sites)


```{r}
# Create filtered datasets for each taxa
dat_acr <- recruit_transect %>% filter(taxa == "Acr")
dat_poc <- recruit_transect %>% filter(taxa == "Poc")
dat_por <- recruit_transect %>% filter(taxa == "Por")
```

```{r}
# Exploratory bar charts 
recruit_transect %>%
  filter(taxa != "Mil") %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = bleaching_status)) +
  geom_col() +
  facet_grid(habitat ~ taxa, scales = "free_y") +  # Not constraint by y-axis
  scale_fill_manual(values = c(Pre = "dodgerblue", Post = "red"),
                    name = "Period") +
  scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
  theme_bw() +
  labs(title = "Recruitment by Habitat and Taxa",
       x = "Year", y = "Recruits per m²") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# recruit_transect %>%
#   filter(taxa != "Mil") %>%
#   mutate(site_habitat = paste(site, habitat, sep = "-")) %>%
#   ggplot(aes(x = factor(year), y = recruits_std, fill = bleaching_status)) +
#   geom_col() +
#   facet_grid(site_habitat ~ taxa) +
#   scale_fill_manual(values = c(Pre = "dodgerblue", Post = "red"),
#                     name = "Period") +
#   theme_bw() +
#   labs(title = "Recruitment by site-habitat combination and taxa",
#        x = "Year", y = "Recruits per m²") + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) 
```

## Model 1: Habitat level recruitment through time
Question: How does coral recruitment differ between lagoon (BR) and forereef (OR) habitats across years from 2013-2024? Did those habitat differences change over time?

-   **Formula: habitat \* year + (1\|site/transect_id)**
-   **site/transect_id: sites pooled as random... island-wide habitat question**
-   **year crossed random effect: absorbs annual variation**


**Get estimate on link scale and then get se since only on link scale and then convert to response space 


How did recruitment differ in the lagoon and forereef between 2013 and 2024?  Recruitment... most abundant genera of corals (Por, Poc, Acr)... How you did it? (Poc is for 1 m squared of forereef) 


### Acropora
Function of separate habitats (do NOT )
** One for lagoon, or for forereef,
THEN look at total recruitment per year without habitat 

```{r}
#| collapse: true
# q1_acr_year <- glmmTMB(
#   recruits_std ~ habitat*factor(year) + (1 | transect_id),
#   family    = tweedie(link = "log"),
#   data      = dat_acr, 
#   na.action = na.exclude
# )
# 
# # Island-wide Acropora 
#  q1_acr_year_island_wide <- glmmTMB(
#   recruits_std ~ factor(year) + (1 | transect_id),
#   family    = tweedie(link = "log"),
#   data      = dat_acr,
#   na.action = na.exclude
# )

# 
# q1_acr_BR <- glmmTMB(
#   recruits_std ~ factor(year) + (1 | transect_id),
#   family    = tweedie(link = "log"),
#   data      = dat_acr %>% filter(habitat == "BR"),
#   na.action = na.exclude
# )
# 
# summary(q1_acr_BR) 
# 
# 
# q1_acr_OR <- glmmTMB(
#   recruits_std ~ factor(year) + (1 | transect_id),
#   family    = tweedie(link = "log"),
#   data      = dat_acr %>% filter(habitat == "OR"),
#   na.action = na.exclude
# )
# 
# summary(q1_acr_OR) 
# 
```


```{r}
# Acr
dat_acr %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  + 
    labs(title = "Habitat-wide Coral Recruitment of Acropora (2013-2024)", caption = "2017 not sampled.") + theme(legend.position = "none")


dat_por %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
    labs(title = "Habitat-wide Coral Recruitment of Porites (2013-2024)", caption = "2017 not sampled.") + theme(legend.position = "none")


dat_poc %>% group_by(habitat, year) %>% 
  summarize(total_recruits = sum(recruits_std)) %>%
  ggplot(aes(x = factor(year), y = total_recruits, fill = habitat)) + geom_col() + facet_wrap(~habitat) + 
    theme(legend.position = "none") + theme_bw() +  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))  +
    labs(title = "Habitat-wide Coral Recruitment of Pocillopera (2013-2024)", caption = "2017 not sampled.") 




```


### Pocillopera

```{r}
# Poc
q1_poc_year <- glmmTMB(
  recruits_std ~ habitat * factor(year) + (1 | transect_id),
  family    = tweedie(link = "log"),
  data      = dat_poc,
  na.action = na.exclude
)

summary(q1_poc_year) 
```

Interpretation: Pre-bleaching OR recruited more than BR. Post-bleaching had flipped pattern: OR dropped sharply while BR held (interaction is p \< .05). OR recruitment seemed to drop significantly post bleaching.


```{r}
#| message: false
#| warning: false
# Visualization
# Observed data (keep all years including 2017 and get one mean observation for habitat-year)
poc_avg_recruits_year <- dat_poc %>% 
  group_by(habitat, year) %>% 
  summarize(avg_recruit = mean(recruits_std, na.rm = TRUE)) %>% 
  ungroup()

# Predictions (2017 excluded) 
poc_pred_grid_year <- dat_poc %>%
  filter(year != 2017) %>%
  distinct(habitat, year)

poc_pred_grid_year$predicted <- predict(q1_poc_year, 
                                        newdata = poc_pred_grid_year,
                                        type = "response",
                                        re.form = NA)

# Join observed to predictions
poc_plot_data <- poc_avg_recruits_year %>%
  left_join(poc_pred_grid_year, by = c("habitat", "year"))

# Plot
ggplot(poc_plot_data) + 
  geom_col(aes(x = factor(year), y = avg_recruit, fill = habitat), 
           alpha = 0.5) +
  geom_point(aes(x = factor(year), y = predicted, col = habitat)) + 
  geom_line(aes(x = factor(year), y = predicted, col = habitat, 
                group = habitat)) +  
  facet_wrap(~ habitat) +
  scale_fill_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),  
    name   = "Observed"
  ) +
  scale_color_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"), 
    name   = "Predicted"
  ) +
  scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
  theme_bw() + 
  theme( # Specify mean observed and predicted recruits
    axis.text.x   = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    title   = "Habitat-wide Coral Recruitment of Pocillopora (2013-2024)",
    y       = "Recruits per m²",
    x       = "Year",
    caption = "Bars = observed mean, points = model predicted. 2017 not sampled."
  )
```

#### Model Validation
```{r}
# Model validation 
## Model validation and assumptions 


# 1. for negative binomial, we will need to test for the overdispersion (qqplot), heteroscedacity, and zero-inflation

# 2. Residual diagnostics (outliers, non-normality, or deviations)
# Plot of predicted vs actual, should not see any curvature, will get noise but on the line mostly 

# Q2A (data never really matches the model... how different is the model from the assumption)
# switch to tweedie model for #Q2a.. try tweedie() 
```


### Porites

```{r}
# Por 
q1_por_year <- glmmTMB(
  recruits_std ~ habitat * factor(year) + (1 | transect_id),
  family    = tweedie(link = "log"),
  data      = dat_por,
  na.action = na.exclude
)

summary(q1_por_year)
```

```{r}
# Visualization
# Observed data (keep all years including 2017)
por_avg_recruits_year <- dat_por %>% 
  group_by(habitat, year) %>% 
  summarize(avg_recruit = mean(recruits_std, na.rm = TRUE)) %>% 
  ungroup()

# Predictions (2017 excluded) 
por_pred_grid_year <- dat_por %>%
  filter(year != 2017) %>%
  distinct(habitat, year)

por_pred_grid_year$predicted <- predict(q1_por_year, 
                                        newdata = por_pred_grid_year,
                                        type = "response",
                                        re.form = NA)

# Join observed to predictions
por_plot_data <- por_avg_recruits_year %>%
  left_join(por_pred_grid_year, by = c("habitat", "year"))

# Plot
ggplot(por_plot_data) + 
  geom_col(aes(x = factor(year), y = avg_recruit, fill = habitat), 
           alpha = 0.5) +
  geom_point(aes(x = factor(year), y = predicted, col = habitat)) + 
  geom_line(aes(x = factor(year), y = predicted, col = habitat, 
                group = habitat)) +  
  facet_wrap(~ habitat) +
  scale_fill_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"),  
    name   = "Observed"
  ) +
  scale_color_manual(
    values = c("Back Reef" = "dodgerblue", "Fore Reef" = "darkorange"), 
    name   = "Predicted"
  ) +
  scale_x_discrete(breaks = as.character(seq(2013, 2024, by = 2))) +
  theme_bw() + 
  theme( # Specify mean observed and predicted recruits
    axis.text.x   = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    title   = "Habitat-wide Coral Recruitment of Porites (2013-2024)",
    y       = "Recruits per m²",
    x       = "Year",
    caption = "Bars = observed mean, points = model predicted. 2017 not sampled."
  )
```


Opposite pattern seen: BR recruited much more than OR pre-bleaching (habitatOR = -1.91 and is significant). Both habitats dropped post bleaching (pre_post is significant); however, the gap is not significant since interaction is not significant.


***offshore dominated by pocillopera because of disturnabce in 2010 (corals came back! Poc is opp)

Bleaching hit recruitment hard across all three taxa and both habitats. Acr and Poc follow the same habitat preference (OR \> BR pre-bleaching, both crash post-bleaching). Por is the differing one; it's a back reef recruiter that was devastated. Wide CI on Porites BR Pre reflects genuine high variability in that cell.


#### Model Validation


## Model 2: Island-wide recruitment by site x habitat before and after bleaching 
Do site × habitat differences in recruitment vary between pre and post-bleaching periods?

Formula: site \* habitat \* bleaching_status + (1\|transect_id) + (1\|year)

### Acropora

```{r}
# Try doing separately for each habitat
q2a_acr <- glmmTMB(
  recruits_std ~ site + habitat + bleaching_status +
                 site:habitat +
                 habitat:bleaching_status +
                 site:bleaching_status +
                 (1 | transect_id) + (1 | year),
  family    = tweedie(),
  data      = dat_acr,
  na.action = na.exclude
)
```

### Pocillopora

```{r}
q2a_poc <- glmmTMB(
  recruits_std ~ site * habitat * bleaching_status + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_poc,
  na.action = na.exclude
)
summary(q2a_poc)

# predictions to get standard errors, conversion to response space!!! 

```


```{r}
# Prediction grid
poc_pred_grid_q2 <- dat_poc %>%
  distinct(site, habitat, bleaching_status)

poc_fit_q2 <- predict(q2a_poc, newdata = poc_pred_grid_q2,
                      type = "response", re.form = NA, se.fit = TRUE)

poc_pred_grid_q2$predicted <- poc_fit_q2$fit
poc_pred_grid_q2$se <- poc_fit_q2$se.fit

# Observed means 

poc_obs_q2 <- dat_poc %>%
  group_by(site, habitat, bleaching_status) %>%
  summarize(observed = mean(recruits_std, na.rm = TRUE), .groups = "drop")


# Combine 

poc_q2_plot_df <- poc_pred_grid_q2 %>%
  left_join(poc_obs_q2, by = c("site", "habitat", "bleaching_status")) %>%
  pivot_longer(cols      = c(predicted, observed),
               names_to  = "type",
               values_to = "recruits") %>%
  mutate(
    bleaching_status = factor(bleaching_status, levels = c("Pre", "Post")),
    type       = factor(type, levels = c("observed", "predicted")),
    site       = factor(site, levels = c("LTER1", "LTER2", "LTER4", "LTER5"))
  )

```

```{r}
# Predictions
ggplot() +
  # Observed bars
  geom_col(data = poc_obs_q2,
           aes(x = habitat, y = observed, fill = bleaching_status),
           position = position_dodge(width = 0.75),
           width = 0.65, colour = "white", linewidth = 0.3) +
  # Predicted points
  geom_point(data = poc_pred_grid_q2,
             aes(x = habitat, y = predicted, group = bleaching_status),
             position = position_dodge(width = 0.75),
             size = 3, shape = 18) +  # shape 18 = filled diamond
  # Predicted ci
  # geom_errorbar(
  #   data = poc_pred_grid_q2,
  #   aes(x    = habitat,
  #       ymin = exp(log(predicted) - 1.96 * (se / predicted)),
  #       ymax = exp(log(predicted) + 1.96 * (se / predicted)),
  #       group = bleaching_status),
  #   position = position_dodge(width = 0.75),
  #   width = 0.25, linewidth = 0.6
  # ) +
  facet_wrap(~ site, ncol = 2) +
  scale_fill_manual(
    values = c(Pre = "#4e9af1", Post = "#f17a4e"),
    name   = "Period",
    labels = c(Pre = "Pre-bleaching (2013-2018)",
               Post = "Post-bleaching (2019-2024)")
  ) +
  labs(
    title    = "Pocillopora recruitment by site x habitat: predicted vs observed",
    subtitle = "Bars = observed mean, point + CI = model predicted (±95% CI)",
    x        = "Habitat",
    y        = "Recruits per m²"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.background   = element_rect(fill = "grey92", colour = NA),
    strip.text         = element_text(face = "bold"),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank(),
    # plot.title         = element_text(face = "bold"),
    # plot.subtitle      = element_text(colour = "grey40")
  )
```

### Porites

```{r}
q2a_por <- glmmTMB(
  recruits_std ~ site * habitat * bleaching_status + (1 | transect_id) + (1 | year),
  family    = tweedie(link = "log"),
  data      = dat_por,
  na.action = na.exclude
)
summary(q2a_por)
```
