---
title: "recruitment_glmm_walkthrough"
format: html
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora between the backreef and forereef before and after the bleaching event?**


## Data Importing and Wrangling 
The following packages were required prior to conducting additional data wrangling: 
```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation 
```


Then, the data (referenced from a separate repository), .

```{r}
# clean_coral <- read_csv(
#   "~/MEDS/capstone/eds-coral-figs-storage/vedika-r-files/data/updated_coral_tidy_2013-2024.csv"
# )

clean_coral <- read_csv(
  "~/MEDS/capstone/eds-coral-data-storage-management/analysisdata/volume_corals.csv"
)
str(clean_coral)


# recruit_raw <- clean_coral %>%
#   filter(!grepl("^P", transect)) %>%
#   dplyr::select(site, habitat, transect, dyn_recruitment, year, taxa) %>%
#   mutate(
#     transect = case_when(
#       transect == "T02_OLD"     ~ "T02",
#       transect == "T03_PRE2020" ~ "T03",
#       transect == "T04_OLD"     ~ "T04",
#       transect == "T02_2021"    ~ "T02",
#       transect == "TO1"         ~ "T01",
#       TRUE   ~ transect
#     ),
#     taxa = if_else(taxa == "A", "Acr", taxa)
#   )

```

```{r}
# Summarize standard recruits to site-habitat-transect and year level 
recruit_transect <- clean_coral %>%
  group_by(taxa, site, habitat, transect, year) %>%
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    recruits_std = n_recruits / 5,
    recruits_int = as.integer(round(recruits_std)),
    pre_post_f   = factor(
      case_when(
        year >= 2013 & year <= 2018 ~ "Pre",
        year >= 2019 & year <= 2024 ~ "Post"
      ),
      levels = c("Pre", "Post")
    ),
    transect_id = paste(site, transect, sep = "_"),
    taxa        = factor(taxa),
    site        = relevel(factor(site),    ref = "LTER1"),
    habitat     = relevel(factor(habitat), ref = "BR")
  ) 
# %>%
#   filter(!is.na(pre_post_f))

view(recruit_transect)
# No NAs
```


```{r}
# Create filtered datasets for each taxa
dat_acr <- recruit_transect %>% filter(taxa == "Acr")
dat_poc <- recruit_transect %>% filter(taxa == "Poc")
dat_por <- recruit_transect %>% filter(taxa == "Por")
```


```{r}
# Exploratory bar chart 
recruit_transect %>%
  filter(taxa != "Mil") %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = pre_post_f)) +
  geom_col() +
  facet_grid(habitat ~ taxa) +
  scale_fill_manual(values = c(Pre = "#4e9af1", Post = "#f17a4e"),
                    name = "Period") +
  theme_bw() +
  labs(title = "Recruitment by habitat and taxa",
       x = "Year", y = "Recruits per 5 m²")

recruit_transect %>%
  filter(taxa != "Mil") %>%
  mutate(site_habitat = paste(site, habitat, sep = "-")) %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = pre_post_f)) +
  geom_col() +
  facet_grid(site_habitat ~ taxa) +
  scale_fill_manual(values = c(Pre = "#4e9af1", Post = "#f17a4e"),
                    name = "Period") +
  theme_bw() +
  labs(title = "Recruitment by site-habitat combination and taxa",
       x = "Year", y = "Recruits per 5 m²")


```


```{r}
# prediction grid 


```


## Question 1: Island wide habitat x period 
- **Formula: habitat * pre_post_f + (1|site/transect_id) + (1|year)**
- **site/transect_id: sites pooled as random... island-wide habitat question**
- **year crossed random effect: absorbs annual variation**

### Acropora
```{r}
cat("Q1 Acr ~ habitat * pre_post_f | site/transect_id + year\n")
cat("Family: Poisson\n")

q1_acr <- glmmTMB(
  recruits_int ~ habitat * pre_post_f + (1 | site/transect_id) + (1 | year),
  family    = nbinom2,
  data      = dat_acr,
  na.action = na.exclude
)

summary(q1_acr)
```
Pre-bleaching, OR recruited much more than BR (habitat OR is 3.08 on log scale and significant). The gap shrank post-bleaching. 

### Porites 
```{r}
cat("Q1 Poc ~ habitat * pre_post_f | site/transect_id + year\n")
cat("Family: nbinom2\n")

q1_poc <- glmmTMB(
  recruits_int ~ habitat * pre_post_f + (1 | site/transect_id) + (1 | year),
  family    = nbinom2,
  data      = dat_poc,
  na.action = na.exclude
)

summary(q1_poc)
testDispersion(q1_poc)
```

Pre-bleaching OR recruited more than BR. Post-bleaching had flipped pattern: OR dropped sharply while BR held (interaction is p < .0001). OR recruitment seemed to drop significantly post bleaching. 



```{r}


#### Joaquin's code
# M1 prediction grids — population-level, no random effects
acr_pred_grid_m1 <- growth_acr |>
  distinct(habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_acr$log10_size_t0),
                               max(growth_acr$log10_size_t0),
                               length.out = 100))
acr_pred_grid_m1$pred_log10_t1 <- predict(m1_acr, newdata = acr_pred_grid_m1,
                                           type = "response", re.form = NA)

poc_pred_grid_m1 <- growth_poc |>
  distinct(habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_poc$log10_size_t0),
                               max(growth_poc$log10_size_t0),
                               length.out = 100))
poc_pred_grid_m1$pred_log10_t1 <- predict(m1_poc, newdata = poc_pred_grid_m1,
                                           type = "response", re.form = NA)

por_pred_grid_m1 <- growth_por |>
  distinct(habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_por$log10_size_t0),
                               max(growth_por$log10_size_t0),
                               length.out = 100))
por_pred_grid_m1$pred_log10_t1 <- predict(m1_por, newdata = por_pred_grid_m1,
                                           type = "response", re.form = NA)


## Kylie
coral_por_hab$fit.Surv <- predict(coral_surv_mod, type = "response", re.form = NA)


```

## Model Validation
```{r}
# Model validation 
## Model validation and assumptions 
# 1. for negative binomial, we will need to test for the overdispersion (qqplot), heteroscedacity, and zero-inflation

# 2. Residual diagnostics (outliers, non-normality, or deviations)

# 3. 

```



