---
title: "recruitment_glmm_walkthrough"
format: html
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora between the backreef and forereef before and after the bleaching event?**


## Data Importing and Wrangling 
The following packages were required prior to conducting additional data wrangling: 
```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation 
```


Then, the data (referenced from a separate repository), .

```{r}
clean_coral <- read_csv(
  "~/MEDS/capstone/eds-coral-figs-storage/vedika-r-files/data/updated_coral_tidy_2013-2024.csv"
)

recruit_raw <- clean_coral %>%
  filter(!grepl("^P", transect)) %>%
  dplyr::select(site, habitat, transect, dyn_recruitment, year, taxa) %>%
  mutate(
    transect = case_when(
      transect == "T02_OLD"     ~ "T02",
      transect == "T03_PRE2020" ~ "T03",
      transect == "T04_OLD"     ~ "T04",
      transect == "T02_2021"    ~ "T02",
      transect == "TO1"         ~ "T01",
      TRUE                      ~ transect
    ),
    taxa = if_else(taxa == "A", "Acr", taxa)
  )

```



```{r}
# Create a function to filter for the dataset of each 




```




