---
title: "recruitment_glmm_walkthrough"
format: html
---

# Capstone CoralModel Project: Assessing the Effect of a Bleaching Event on Demographic Rates Variation in Coral Communities in Moorea, French Polynesia

## Introduction

This document details the analysis of coral recruitment for three different taxa based on survey data recorded over nearly a decade by Dr. Hunter Lenihan for the Mo'orea Long Term Ecological Network site (MCR-LTER). The purpose of this analysis is to serve as a supporting walk through for understanding recruitment as one of three demographic profiles impacted by the 2019 bleaching event. The following question was utilized to guide the analysis:

**What are island-wide differences in coral recruitment for Acropora, Porites, and Pocilopora between the backreef and forereef before and after the bleaching event?**


## Data Importing and Wrangling 
The following packages were required prior to conducting additional data wrangling: 
```{r}
library(tidyverse) # For tidy data
library(glmmTMB) # For GLMM model 
library(ggeffects) # For calculating marginal effects of each model 
library(DHARMa) # For model checks and validation 
```


Then, the data (referenced from a separate repository), .

```{r}
clean_coral <- read_csv(
  "~/MEDS/capstone/eds-coral-figs-storage/vedika-r-files/data/updated_coral_tidy_2013-2024.csv"
)

recruit_raw <- clean_coral %>%
  filter(!grepl("^P", transect)) %>%
  dplyr::select(site, habitat, transect, dyn_recruitment, year, taxa) %>%
  mutate(
    transect = case_when(
      transect == "T02_OLD"     ~ "T02",
      transect == "T03_PRE2020" ~ "T03",
      transect == "T04_OLD"     ~ "T04",
      transect == "T02_2021"    ~ "T02",
      transect == "TO1"         ~ "T01",
      TRUE   ~ transect
    ),
    taxa = if_else(taxa == "A", "Acr", taxa)
  )

```

```{r}
# Summarize standard recruits to transect and year level 
recruit_transect <- recruit_raw %>%
  group_by(taxa, site, habitat, transect, year) %>%
  summarize(n_recruits = sum(dyn_recruitment, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    recruits_std = n_recruits / 5,
    recruits_int = as.integer(round(recruits_std)),
    pre_post_f   = factor(
      case_when(
        year >= 2013 & year <= 2018 ~ "Pre",
        year >= 2019 & year <= 2024 ~ "Post"
      ),
      levels = c("Pre", "Post")
    ),
    transect_id = paste(site, transect, sep = "_"),
    taxa        = factor(taxa),
    site        = relevel(factor(site),    ref = "LTER1"),
    habitat     = relevel(factor(habitat), ref = "BR")
  ) %>%
  filter(!is.na(pre_post_f))


```


```{r}
# Create filtered datasets for each taxa
dat_acr      <- recruit_transect %>% filter(taxa == "Acr")
dat_poc      <- recruit_transect %>% filter(taxa == "Poc")
dat_por      <- recruit_transect %>% filter(taxa == "Por")
```


```{r}
# Exploratory bar chart 
recruit_transect %>%
  filter(taxa != "Mil") %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = pre_post_f)) +
  geom_col() +
  facet_grid(habitat ~ taxa) +
  scale_fill_manual(values = c(Pre = "#4e9af1", Post = "#f17a4e"),
                    name = "Period") +
  theme_bw() +
  labs(title = "Recruitment by habitat and taxa",
       x = "Year", y = "Recruits per 5 m²")

recruit_transect %>%
  filter(taxa != "Mil") %>%
  mutate(site_habitat = paste(site, habitat, sep = "-")) %>%
  ggplot(aes(x = factor(year), y = recruits_std, fill = pre_post_f)) +
  geom_col() +
  facet_grid(site_habitat ~ taxa) +
  scale_fill_manual(values = c(Pre = "#4e9af1", Post = "#f17a4e"),
                    name = "Period") +
  theme_bw() +
  labs(title = "Recruitment by site-habitat combination and taxa",
       x = "Year", y = "Recruits per 5 m²")


```


```{r}
library(tidyverse)
library(ggridges)
library(patchwork)

# ── 1. Ridge plot: recruitment distribution by year × taxa ──────────────────
p_ridge <- ggplot(recruit_transect,
                  aes(x     = recruits_std,
                      y     = factor(year),
                      fill  = taxa,
                      color = taxa)) +
  geom_density_ridges(alpha = 0.55, scale = 1.2,
                      quantile_lines = TRUE, quantiles = 2) +   # median line
  facet_wrap(~ taxa, ncol = 3) +
  scale_fill_brewer(palette  = "Set2", guide = "none") +
  scale_color_brewer(palette = "Set2", guide = "none") +
  labs(title    = "Recruitment Distribution by Year & Taxa",
       subtitle = "Standardised recruits per transect  |  median shown",
       x = "Recruits (std)",
       y = NULL) +
  theme_ridges(grid = TRUE) +
  theme(strip.text = element_text(face = "bold"))

# ── 2. Box plot: per-taxa distribution collapsed across years ───────────────
p_box <- ggplot(recruit_transect,
                aes(x    = taxa,
                    y    = recruits_std,
                    fill = taxa)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5,
               outlier.alpha = 0.5, width = 0.55) +
  geom_jitter(aes(color = taxa), width = 0.15,
              alpha = 0.25, size = 0.8) +
  scale_fill_brewer(palette  = "Set2", guide = "none") +
  scale_color_brewer(palette = "Set2", guide = "none") +
  labs(title    = "Recruitment Distribution by Taxa",
       subtitle = "Standardised recruits per transect",
       x = NULL,
       y = "Recruits (std)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        axis.text.x = element_text(face = "bold"))

# ── 3. Combine ──────────────────────────────────────────────────────────────
p_ridge
p_box 

```




```{r}

```







