---
title: "Coral Growth — Size-Dependent Growth Rate Analysis"
subtitle: "GLMM analysis of proportional growth rates by taxa, site, habitat, and bleaching period"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
editor: visual
---

# Overview

This document models **size-dependent proportional growth rates** for three coral taxa (*Acropora*, *Pocillopora*, *Porites*) using generalized linear mixed models (GLMMs) fitted with `glmmTMB`.

Two research questions are addressed:

1.  **Model 1 (Habitat):** Does size-dependent growth rate differ between forereef and backreef habitats around the island before and after the 2019 bleaching event?
2.  **Model 2 (Site × Habitat):** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 backreef, LTER1 forereef, LTER2 backreef, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Reference group for all models:** backreef (Model 1) or LTER1 backreef (Model 2), before bleaching.

All models use `log(size_t1)` as the response variable, with `log(size_t0)` as a covariate. The coefficient on `log_size_t0` captures the size-dependence of growth.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(glmmTMB)
library(performance)
library(DHARMa)
library(emmeans)
library(dplyr)
library(purrr)
library(gt)
```

------------------------------------------------------------------------

# Data Preparation

## Load & Clean

```{r}
#| label: load-clean
#| message: false
#| warning: false

coral_demography_analysis_data <- read_csv(here::here("..", "eds-coral-data-storage-management", "analysisdata", "coral_demography_analysis_data.csv"), guess_max = 10000) # Correct column type 

```

## Filter & Model Preparation

```{r}
#| label: filter-data

# Exclusion criteria:
# - Millepora excluded (insufficient sample size)
# - Require both size measurements (t0 and t1 non-missing)
# - Only consecutive year pairs (year_t1 - year == 1)
# - Remove zero/negative volumes (measurement errors)
growth_data <- coral_demography_analysis_data %>%
  mutate(year_t1 = year + 1) %>% 
  filter(
    taxa != "Mil",
    !is.na(size_t0), 
    !is.na(size_t1),
    !is.na(year_t1), 
    year_t1 - year == 1,  # consecutive years only
    size_t0 > 0, size_t1 > 0) %>%
  mutate(
    # Set reference levels
    taxa        = relevel(factor(taxa),    ref = "Acr"),
    habitat     = relevel(factor(habitat), ref = "BR"),
    site        = relevel(factor(site),    ref = "LTER1"),
    period = factor(period), 
    # Log-transform size variables
    log10_size_t0 = log10(size_t0),
    log10_size_t1 = log10(size_t1)
  )

# Split by taxa for separate models
growth_acr <- growth_data %>% filter(taxa == "Acr")
growth_poc <- growth_data %>% filter(taxa == "Poc")
growth_por <- growth_data %>% filter(taxa == "Por")
```

------------------------------------------------------------------------

# Model 1: Habitat-Level Growth Before/After Bleaching

**Question:** Does size-dependent growth rate differ between forereef and backreef habitats before and after the 2019 bleaching event?

**Fixed effects:** `log_size_t0 * habitat * period`

**Random effects:** Colonies nested in transects nested in sites `(1 | site/transect_id/coral_id)` + inter-annual variation `(1 | year)`

**Reference group:** Backreef, before bleaching.

## Fit Models

```{r}
#| label: m1-fit
#| message: false
#| warning: false

m1_acr <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | transect_id/coral_id) + (1 | year),
  family = gaussian(),
  data   = growth_acr
)
m1_poc <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | transect_id/coral_id) + (1 | year),
  family = gaussian(),
  data   = growth_poc
)

m1_por <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | transect_id/coral_id) + (1 | year),
  family = gaussian(),
  data   = growth_por
)
```

## Model Summaries

```{r}
#| label: m1-summaries

summary(m1_acr)
summary(m1_poc)
summary(m1_por)
```

```{r}
# Acropora
emtrends(m1_acr, ~ habitat * period, var = "log10_size_t0")
contrast(emtrends(m1_acr, ~ habitat * period, var = "log10_size_t0"),
         method = "pairwise", by = "habitat")

# Pocillopora
emtrends(m1_poc, ~ habitat * period, var = "log10_size_t0")
contrast(emtrends(m1_poc, ~ habitat * period, var = "log10_size_t0"),
         method = "pairwise", by = "habitat")

# Porites
emtrends(m1_por, ~ habitat * period, var = "log10_size_t0")
contrast(emtrends(m1_por, ~ habitat * period, var = "log10_size_t0"),
         method = "pairwise", by = "habitat")
```

```{r}
models <- list(ACR = m1_acr, POC = m1_poc, POR = m1_por)

# Group-specific slopes
slopes <- imap_dfr(models, ~ as.data.frame(
  emtrends(.x, ~ habitat * period, var = "log10_size_t0")
) |> mutate(taxon = .y))

# Within-habitat period contrasts
contrasts <- imap_dfr(models, ~ as.data.frame(
  pairs(emtrends(.x, ~ period | habitat, var = "log10_size_t0"))
) |> mutate(taxon = .y))

# Clean table
contrasts |>
  select(taxon, habitat, contrast, estimate, SE, z.ratio, p.value) |>
  mutate(
    across(c(estimate, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) |>
  arrange(taxon, habitat) |>
  gt(groupname_col = "taxon") |>
  cols_label(
    habitat = "Habitat",
    contrast = "Contrast",
    estimate = "Δ slope",
    z.ratio = "z",
    p.value = "p",
    sig = ""
  ) |>
  tab_header(title = "Change in size-dependent growth slope after 2019 bleaching")
```

```{r}
#| label: m1-predictions
# M1 prediction grids — population-level, no random effects

# Build a prediction grid: every habitat × period combination paired with
# 100 evenly-spaced log10_size_t0 values spanning the observed range.
acr_pred_grid_m1 <- growth_acr %>%
  distinct(habitat, period) %>%  # unique habitat × period combos (4 rows: BR/OR × 0/1)
  crossing(log10_size_t0 = seq(min(growth_acr$log10_size_t0), # full crossing : 4 × 100 = 400 rows
                               max(growth_acr$log10_size_t0),
                               length.out = 100))

acr_pred_grid_m1$pred_log10_t1 <- predict(m1_acr, newdata = acr_pred_grid_m1,
                                          type = "response", re.form = NA)
poc_pred_grid_m1 <- growth_poc %>%
  distinct(habitat, period) %>%
  crossing(log10_size_t0 = seq(min(growth_poc$log10_size_t0),
                               max(growth_poc$log10_size_t0),
                               length.out = 100))
poc_pred_grid_m1$pred_log10_t1 <- predict(m1_poc, newdata = poc_pred_grid_m1,
                                           type = "response", re.form = NA)

por_pred_grid_m1 <- growth_por %>%
  distinct(habitat, period) %>%
  crossing(log10_size_t0 = seq(min(growth_por$log10_size_t0),
                               max(growth_por$log10_size_t0),
                               length.out = 100))
por_pred_grid_m1$pred_log10_t1 <- predict(m1_por, newdata = por_pred_grid_m1,
                                           type = "response", re.form = NA)

```

```{r}
#| label: m1-plots
ggplot() +
  geom_point(data = growth_acr,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  geom_line(data = acr_pred_grid_m1,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw() +
  labs(title = "Acropora Size-Dependent Growth Across Habitats", x = "Initial size (cm³)", y = "Final size (cm³)", color = "Period")

ggplot() +
  geom_point(data = growth_poc,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = poc_pred_grid_m1,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw() +
  labs(title = "Pocillopora Size-Dependent Growth Across Habitats", x = "Initial size (cm³)", y = "Final size (cm³)", color = "Period")

ggplot() +
  geom_point(data = growth_por,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = por_pred_grid_m1,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Pre-bleaching", "1" = "Post-bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw() +
  labs(title = "Porites Size-Dependent Growth Across Habitats", x = "Initial size (cm³)", y = "Final size (cm³)", color = "Period")

ggsave(here::here("growth", "figs", "acr_growth_hab_plot.png"), height = 5, width = 10)
ggsave(here::here("growth", "figs", "poc_growth_hab_plot.png"), height = 5, width = 10)
ggsave(here::here("growth", "figs", "por_growth_hab_plot.png"), height = 5, width = 10)
   
```

## Model Checks

### *Acropora* residuals

```{r}
acr_growth_mod_output <- simulateResiduals(fittedModel = m1_acr, plot = FALSE)
plot(acr_growth_mod_output)
```

### *Pocillopora* residuals

```{r}
poc_growth_mod_output <- simulateResiduals(fittedModel = m1_poc, plot = FALSE)
plot(poc_growth_mod_output)
```

### *Porites* residuals

```{r}
por_growth_mod_output <- simulateResiduals(fittedModel = m1_por, plot = FALSE)
plot(por_growth_mod_output)
```

# Model 2: Island-Wide Growth by Site × Habitat Before/After Bleaching

**Question:** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 backreef, LTER1 forereef, LTER2 backreef, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Fixed effects:** `log_size_t0 * site * habitat * period`

**Random effects:** Colonies nested in transects `(1 | transect_id/coral_id)` + inter-annual variation `(1 | year)`

Note: Site is included as a **fixed effect** here because the question explicitly compares specific named sites. Transect nesting is simplified to `transect_id/coral_id` since site is already captured in the fixed effects and in `transect_id`.

**Reference group:** LTER1 backreef, before bleaching.

## Fit Models

```{r}
#| label: m2-fit
#| message: false
#| warning: false

m2_acr <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_id) + (1 | year),
  family = gaussian(),
  data   = growth_acr
)

m2_poc <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_id) + (1 | year),
  family = gaussian(),
  data   = growth_poc
)

m2_por <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_id) + (1 | year),
  family = gaussian(),
  data   = growth_por
)
```

## Model Summaries

```{r}
#| label: m2-summaries

summary(m2_acr)
summary(m2_poc)
summary(m2_por)
```

```{r}
models2 <- list(ACR = m2_acr, POC = m2_poc, POR = m2_por)

# Group-specific slopes (by site × habitat × period)
slopes2 <- imap_dfr(models2, ~ as.data.frame(
  emtrends(.x, ~ site * habitat * period, var = "log10_size_t0")
) |> mutate(taxon = .y))

# Within-habitat period contrasts at each site
contrasts2 <- imap_dfr(models2, ~ as.data.frame(
  pairs(emtrends(.x, ~ period | site * habitat, var = "log10_size_t0"))
) |> mutate(taxon = .y))

# Clean table
contrasts2 |>
  select(taxon, site, habitat, contrast, estimate, SE, z.ratio, p.value) |>
  mutate(
    across(c(estimate, SE, z.ratio), \(x) round(x, 3)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    p.value = format.pval(p.value, digits = 2, eps = 0.001)
  ) |>
  arrange(taxon, site, habitat) |>
  gt(groupname_col = "taxon") |>
  cols_label(
    site = "Site",
    habitat = "Habitat",
    contrast = "Contrast",
    estimate = "Δ slope",
    z.ratio = "z",
    p.value = "p",
    sig = ""
  ) |>
  tab_header(title = "Change in size-dependent growth slope after 2019 bleaching, by site")
```

```{r}
#| label: m2-predictions
acr_pred_grid_m2 <- growth_acr %>%
  distinct(site, habitat, period) %>%
  crossing(log10_size_t0 = seq(min(growth_acr$log10_size_t0),
                               max(growth_acr$log10_size_t0),
                               length.out = 100))
acr_pred_grid_m2$pred_log10_t1 <- predict(m2_acr, newdata = acr_pred_grid_m2,
                                           type = "response", re.form = NA)

poc_pred_grid_m2 <- growth_poc %>%
  distinct(site, habitat, period) %>%
  crossing(log10_size_t0 = seq(min(growth_poc$log10_size_t0),
                               max(growth_poc$log10_size_t0),
                               length.out = 100))
poc_pred_grid_m2$pred_log10_t1 <- predict(m2_poc, newdata = poc_pred_grid_m2,
                                           type = "response", re.form = NA)

por_pred_grid_m2 <- growth_por %>%
  distinct(site, habitat, period) %>%
  crossing(log10_size_t0 = seq(min(growth_por$log10_size_t0),
                               max(growth_por$log10_size_t0),
                               length.out = 100))
por_pred_grid_m2$pred_log10_t1 <- predict(m2_por, newdata = por_pred_grid_m2,
                                           type = "response", re.form = NA)
```

```{r}
#| label: m2-plots
ggplot() +
  geom_point(data = growth_acr,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = acr_pred_grid_m2,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period)) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  labs(x = "Initial size (cm³)",
       y = "Final size (cm³)", 
       title = "Acropora Size-Dependent Growth Across Site and Habitat", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()

ggplot() +
  geom_point(data = growth_poc,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = poc_pred_grid_m2,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period)) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  labs(x = "Initial size (cm³)", 
       y = "Final size (cm³)", 
       title = "Pocillopora Size-Dependent Growth Across Site and Habitat", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()

ggplot() +
  geom_point(data = growth_por,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = por_pred_grid_m2,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period)) +
  geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Lagoon", "OR" = "Forereef"))) +
  labs(x = "Initial size (cm³)", 
       y = "Final size (cm³)", 
       title = "Porites Size-Dependent Growth Across Site and Habitat", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()


ggsave(here::here("growth", "figs", "acr_site_growth_plot.png"), height = 5, width = 10)
ggsave(here::here("growth", "figs", "poc_site_growth_plot.png"), height = 5, width = 10)
ggsave(here::here("growth", "figs", "por_site_growth_plot.png"), height = 5, width = 10)
```

## Model Checks

### *Acropora* residuals

```{r}
acr_growth_mod2_output <- simulateResiduals(fittedModel = m2_acr, plot = FALSE)
plot(acr_growth_mod2_output)
```

### *Pocillopora* residuals

```{r}
poc_growth_mod2_output <- simulateResiduals(fittedModel = m2_poc, plot = FALSE)
plot(poc_growth_mod2_output)
```

### *Porites* residuals

```{r}
por_growth_mod2_output <- simulateResiduals(fittedModel = m2_por, plot = FALSE)
plot(por_growth_mod2_output)
```
