---
title: "Coral Growth — Size-Dependent Growth Rate Analysis"
subtitle: "GLMM analysis of proportional growth rates by taxa, site, habitat, and bleaching period"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
editor: visual
---

# Overview

This document models **size-dependent proportional growth rates** for three coral taxa (*Acropora*, *Pocillopora*, *Porites*) using generalized linear mixed models (GLMMs) fitted with `glmmTMB`.

Two research questions are addressed:

1.  **Model 1 (Habitat):** Does size-dependent growth rate differ between forereef and backreef habitats around the island before and after the 2019 bleaching event?
2.  **Model 2 (Site × Habitat):** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 backreef, LTER1 forereef, LTER2 backreef, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Reference group for all models:** backreef (Model 1) or LTER1 backreef (Model 2), before bleaching.

All models use `log(size_t1)` as the response variable, with `log(size_t0)` as a covariate. The coefficient on `log_size_t0` captures the size-dependence of growth.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(glmmTMB)
```

------------------------------------------------------------------------

# Data Preparation

## Load & Clean

```{r}
#| label: load-clean
#| message: false
#| warning: false

volume_coral <- read_csv(here::here("..", "eds-coral-data-storage-management", "data", "volume_corals.csv"))

glimpse(volume_coral)
```

For each coral, look at its years in order and check if there's ever a jump of more than 1 year between observations. If so, keep all rows for that coral. The result is a dataset of only the corals that "disappeared" for at least one survey year and reappeared later - this would indicate recruitment of the same species at the exact location where another had died.

```{r}
# For each coral_number, check if years are non-consecutive (gap = died & same species recruited)
gaps <- volume_coral |>
  arrange(coral_number, year) |>
  group_by(coral_number) |>
  filter(any(diff(year) > 1)) |> # any() Given set of logical vectors, is at least one value true? 
  ungroup()                      # diff() calculates difference between years 
  

# total number of instances where this took place 
n_distinct(gaps$coral_number)

```

```{r}
# Flag coral_numbers where size drops dramatically (e.g., large coral -> tiny recruit)
resets <- volume_coral |>
  select(year, site, habitat, transect_id, coral_number, size_t0, size_t1) %>% 
  arrange(coral_number, year) |>
  group_by(coral_number) |>
  mutate(pct_change = (size_t1 - size_t0) / size_t0) |>
  filter(pct_change < -0.95) |>
  ungroup()

reset_ids <- resets |> pull(coral_number) |> unique()

resets <- volume_coral |>
  filter(coral_number %in% reset_ids) |>
  select(year, coral_number, size_t0, size_t1, dyn_death, dyn_recruitment) |>
  arrange(coral_number, year)
```

```{r}
volume_coral <- volume_coral |>
  arrange(coral_number, year) |>
  group_by(coral_number) |>
  mutate(year_t1 = lead(year)) |>
  ungroup()
```

## Filter & Model Preparation

```{r}
#| label: filter-data

# Exclusion criteria:
# - Millepora excluded (insufficient sample size)
# - Require both size measurements (t0 and t1 non-missing)
# - Only consecutive year pairs (year_t1 - year == 1)
# - Remove zero/negative volumes (measurement errors)
# - Remove implausible growth ratios: >10x or <0.1x annual change
#   (confirmed as measurement error — spread across all years/taxa,
#    not concentrated around 2019 bleaching event)
growth_data <- volume_coral |>
  filter(
    taxa != "Mil",
    !is.na(size_t0), 
    !is.na(size_t1),
    !is.na(year_t1), 
    year_t1 - year == 1,  # consecutive years only
    size_t0 > 0, size_t1 > 0) |>
  mutate(
    # Set reference levels
    taxa        = relevel(factor(taxa),    ref = "Acr"),
    habitat     = relevel(factor(habitat), ref = "BR"),
    site        = relevel(factor(site),    ref = "LTER1"),
    period = factor(period), 
    # Log-transform size variables
    log10_size_t0 = log10(size_t0),
    log10_size_t1 = log10(size_t1)
  )

# Split by taxa for separate models
growth_acr <- growth_data |> filter(taxa == "Acr")
growth_poc <- growth_data |> filter(taxa == "Poc")
growth_por <- growth_data |> filter(taxa == "Por")

# Check the range of growth ratios now in por
growth_por |>
  mutate(growth_ratio = size_t1 / size_t0) |>
  summarise(min = min(growth_ratio), max = max(growth_ratio),
            n_extreme = sum(growth_ratio > 10 | growth_ratio < 0.1))

# Are the extremes concentrated in a particular year/site/habitat?
growth_por |>
  mutate(growth_ratio = size_t1 / size_t0) |>
  filter(growth_ratio > 10 | growth_ratio < 0.1) |>
  count(year, site, habitat)
```

------------------------------------------------------------------------

# Model 1: Habitat-Level Growth Before/After Bleaching

**Question:** Does size-dependent growth rate differ between forereef and backreef habitats before and after the 2019 bleaching event?

**Fixed effects:** `log_size_t0 * habitat * period`

**Random effects:** Colonies nested in transects nested in sites `(1 | site/transect_id/coral_number)` + inter-annual variation `(1 | year)`

**Reference group:** Backreef, before bleaching.

## Fit Models

```{r}
#| label: m1-fit
#| message: false
#| warning: false

m1_acr <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | site/transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_acr
)

m1_poc <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | site/transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_poc
)

m1_por <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * habitat * period +
    (1 | site/transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_por
)
```

## Model Summaries

```{r}
#| label: m1-summaries

summary(m1_acr)
summary(m1_poc)
summary(m1_por)
```

------------------------------------------------------------------------

```{r}
#| label: m1-predictions
# M1 prediction grids — population-level, no random effects
acr_pred_grid_m1 <- growth_acr |>
  distinct(habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_acr$log10_size_t0),
                               max(growth_acr$log10_size_t0),
                               length.out = 100))
acr_pred_grid_m1$pred_log10_t1 <- predict(m1_acr, newdata = acr_pred_grid_m1,
                                           type = "response", re.form = NA)

poc_pred_grid_m1 <- growth_poc |>
  distinct(habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_poc$log10_size_t0),
                               max(growth_poc$log10_size_t0),
                               length.out = 100))
poc_pred_grid_m1$pred_log10_t1 <- predict(m1_poc, newdata = poc_pred_grid_m1,
                                           type = "response", re.form = NA)

por_pred_grid_m1 <- growth_por |>
  distinct(habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_por$log10_size_t0),
                               max(growth_por$log10_size_t0),
                               length.out = 100))
por_pred_grid_m1$pred_log10_t1 <- predict(m1_por, newdata = por_pred_grid_m1,
                                           type = "response", re.form = NA)

```

```{r}
#| label: m1-plots
ggplot() +
  geom_point(data = growth_acr,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = acr_pred_grid_m1,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw() +
  labs(title = "Acropora Growth", x = "Initial size (cm³)", y = "Final size (cm³)", color = "Period")

ggplot() +
  geom_point(data = growth_poc,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = poc_pred_grid_m1,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw() +
  labs(title = "Pocillopora Growth", x = "Initial size (cm³)", y = "Final size (cm³)", color = "Period")

ggplot() +
  geom_point(data = growth_por,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = por_pred_grid_m1,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period), linewidth = 1) +
  facet_grid(~ habitat, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw() +
  labs(title = "Porites Growth", x = "Initial size (cm³)", y = "Final size (cm³)", color = "Period")

   
```

# Model 2: Island-Wide Growth by Site × Habitat Before/After Bleaching

**Question:** Does size-dependent growth rate differ across all site-habitat combinations (LTER1 backreef, LTER1 forereef, LTER2 backreef, LTER2 forereef, etc.) before and after the 2019 bleaching event, island-wide?

**Fixed effects:** `log_size_t0 * site * habitat * period`

**Random effects:** Colonies nested in transects `(1 | transect_id/coral_number)` + inter-annual variation `(1 | year)`

Note: Site is included as a **fixed effect** here because the question explicitly compares specific named sites. Transect nesting is simplified to `transect_id/coral_number` since site is already captured in the fixed effects and in `transect_id`.

**Reference group:** LTER1 backreef, before bleaching.

## Fit Models

```{r}
#| label: m2-fit
#| message: false
#| warning: false

m2_acr <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_acr
)

m2_poc <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_poc
)

m2_por <- glmmTMB(
  log10_size_t1 ~ log10_size_t0 * site * habitat * period +
    (1 | transect_id/coral_number) + (1 | year),
  family = gaussian(),
  data   = growth_por
)
```

## Model Summaries

```{r}
#| label: m2-summaries

summary(m2_acr)
summary(m2_poc)
summary(m2_por)
```

```{r}
#| label: m2-predictions
acr_pred_grid_m2 <- growth_acr |>
  distinct(site, habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_acr$log10_size_t0),
                               max(growth_acr$log10_size_t0),
                               length.out = 100))
acr_pred_grid_m2$pred_log10_t1 <- predict(m2_acr, newdata = acr_pred_grid_m2,
                                           type = "response", re.form = NA)

poc_pred_grid_m2 <- growth_poc |>
  distinct(site, habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_poc$log10_size_t0),
                               max(growth_poc$log10_size_t0),
                               length.out = 100))
poc_pred_grid_m2$pred_log10_t1 <- predict(m2_poc, newdata = poc_pred_grid_m2,
                                           type = "response", re.form = NA)

por_pred_grid_m2 <- growth_por |>
  distinct(site, habitat, period) |>
  crossing(log10_size_t0 = seq(min(growth_por$log10_size_t0),
                               max(growth_por$log10_size_t0),
                               length.out = 100))
por_pred_grid_m2$pred_log10_t1 <- predict(m2_por, newdata = por_pred_grid_m2,
                                           type = "response", re.form = NA)
```

```{r}
#| label: m2-plots
ggplot() +
  geom_point(data = growth_acr,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = acr_pred_grid_m2,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period)) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  labs(x = "Initial size (cm³)", y = "Final size (cm³)", title = "Acropora Growth", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()

ggplot() +
  geom_point(data = growth_poc,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = poc_pred_grid_m2,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period)) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  labs(x = "Initial size (cm³)", y = "Final size (cm³)", title = "Pocillopora Growth", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()

ggplot() +
  geom_point(data = growth_por,
             aes(x = log10_size_t0, y = log10_size_t1), alpha = 0.3, size = 0.8) +
  geom_line(data = por_pred_grid_m2,
            aes(x = log10_size_t0, y = pred_log10_t1, color = period)) +
  scale_color_manual(values = c("0" = "dodgerblue", "1" = "red"),
                     labels = c("0" = "Before bleaching", "1" = "After bleaching")) +
  facet_grid(habitat ~ site, labeller = labeller(habitat = c("BR" = "Back Reef", "OR" = "Fore Reef"))) +
  labs(x = "Initial size (cm³)", y = "Final size (cm³)", title = "Porites Growth", color = "Period") +
  scale_x_continuous(labels = scales::label_math(10^.x)) +
  scale_y_continuous(labels = scales::label_math(10^.x)) +
  theme_bw()
```
